---
title: "Actuarial Senescence Analysis"
author: "Luisa Fitzpatrick"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(here)
library(tidyverse)
library(BaSTA)
library(cowplot)
library(ggplot2)
library(gridExtra)
library(lme4)
library(flextable)
library(officer)

# Shared plotting aesthetics
site_colors <- c("Warm lowland" = "coral1", "Cool highland" = "cornflowerblue")

theme_publication <- theme_classic(base_size = 12) +
  theme(
    axis.text = element_text(color = "black"),
    axis.line = element_line(color = "black"),
    legend.position = "none",
    plot.margin = margin(10, 10, 10, 10))
```

# 1. Load and prepare data
```{r load-data}
data <- read.csv(here("data", "processed", "census_data.csv"),
                 na = c("", "NA", ".", "?")) %>%
  select(Year, FemID, Site, Age, DeathYear) %>%
  rename(year = Year, femid = FemID, site = Site, death = DeathYear) %>%
  mutate(
    birth = year - Age,
    femid = as.factor(femid),
    site  = as.factor(site),
    death = as.integer(death),  # NAs introduced here are expected: individuals
    birth = as.integer(birth))  # with no recorded death year are right-censored
                                # (death set to 0 in prepare_basta_input)
data <- data %>% filter(!is.na(femid))
```

# 2. Helper functions
```{r helper-functions}
# Prepare BaSTA input matrix from site-level data.
# birth/death NAs are replaced with 0 (BaSTA convention for unknown dates).
# Row order of birthDeath is matched to Y matrix by ID to avoid misalignment.
prepare_basta_input <- function(site_data, study_start, study_end) {
  site_data$birth[is.na(site_data$birth)] <- 0
  site_data$death[is.na(site_data$death)] <- 0
  
  Y <- CensusToCaptHist(ID = site_data$femid, d = site_data$year, dformat = "yyyy")
  
  birthDeath <- site_data %>%
    select(femid, birth, death) %>%
    distinct(femid, .keep_all = TRUE)
  
  # Match row order to Y matrix rather than assuming sorted order
  birthDeath <- birthDeath[match(Y[, 1], birthDeath$femid), ]
  
  n_year_cols <- ncol(Y) - 1  # minus the ID column
  inputMat <- as.data.frame(cbind(Y, birthDeath[, -1]))
  inputMat <- inputMat[, c(1, n_year_cols + 2, n_year_cols + 3, 2:(n_year_cols + 1))]
  
  cat("Total individuals:", nrow(inputMat), "\n")
  cat("Total capture events:", sum(Y[, -1]), "\n")
  
  DataCheck(inputMat, studyStart = study_start, studyEnd = study_end, silent = FALSE)
  
  return(inputMat)
}
```


# ============================================================
# COOL HIGHLAND (CENTRAL PLATEAU) ANALYSIS
# ============================================================

# 3. Prepare Plateau data
```{r prep-cp}
dataCP <- data %>% filter(site == "CP") %>% select(-site)
inputMat_CP <- prepare_basta_input(dataCP, study_start = 2000, study_end = 2024)
```

# 4. Model selection — compare mortality functions (CP)
```{r basta-cp-multi, eval=FALSE}
set.seed(123)
multiout_CP <- multibasta(
  object = inputMat_CP, studyStart = 2000, studyEnd = 2024,
  models = c("EX", "GO", "WE", "LO"),
  shapes  = c("simple", "Makeham"),
  nsim = 4, niter = 10000, burnin = 2000, thinning = 10, ncpus = 4)

saveRDS(multiout_CP, here("basta_outputs", "multiout_CP.rds"))
```

```{r load-multi-cp}
multiout_CP <- readRDS(here("basta_outputs", "multiout_CP.rds"))
summary(multiout_CP, digits = 3)
# Result: Gompertz-Makeham selected as best model by DIC
```

# 5. Validate recapture assumption (CP)

BaSTA assumes constant recapture probability across years by default. We test
this by comparing constant vs. time-varying recapture under the best-fitting
model (Gompertz-Makeham). Lower DIC supports the better-fitting recapture
structure.

```{r recapture-dic-cp, eval=FALSE}
out_CP_const <- basta(
  object = inputMat_CP, studyStart = 2000, studyEnd = 2024,
  model = "GO", shape = "Makeham", minAge = 2,
  nsim = 4, niter = 50000, burnin = 5001, thinning = 20,
  updateJumps = TRUE)

out_CP_tv <- basta(
  object = inputMat_CP, studyStart = 2000, studyEnd = 2024,
  model = "GO", shape = "Makeham", minAge = 2,
  recaptureType = "year",
  nsim = 4, niter = 50000, burnin = 5001, thinning = 20,
  updateJumps = TRUE)

saveRDS(out_CP_const, here("basta_outputs", "out_CP_const.rds"))
saveRDS(out_CP_tv,    here("basta_outputs", "out_CP_tv.rds"))
```

```{r recapture-dic-cp-results}
out_CP_const <- readRDS(here("basta_outputs", "out_CP_const.rds"))
out_CP_tv    <- readRDS(here("basta_outputs", "out_CP_tv.rds"))

cat("CP — Constant recapture DIC:", out_CP_const$DIC["DIC"], "\n")
cat("CP — Time-varying recapture DIC:", out_CP_tv$DIC["DIC"], "\n")

cat("Constant recapture:\n")
print(out_CP_const$coefficients[1:3, ])

cat("\nTime-varying recapture:\n")
print(out_CP_tv$coefficients[1:3, ])

# Time-varying fits best
```

# 6. Final model — Gompertz-Makeham (CP) with time-varying recapture
```{r basta-cp-final, eval=FALSE}
set.seed(123)
out_CP_GM_final <- basta(
  object = inputMat_CP, studyStart = 2000, studyEnd = 2024,
  model = "GO", shape = "Makeham", minAge = 2,
  recaptureType = "year",
  nsim = 4, niter = 50000, burnin = 5001, thinning = 20,
  updateJumps = TRUE, lifeTable = TRUE)

saveRDS(out_CP_GM_final, here("basta_outputs", "out_CP_GM_final.rds"))
```

```{r cp-results}
out_CP_GM_final <- readRDS(here("basta_outputs", "out_CP_GM_final.rds"))
summary(out_CP_GM_final, digits = 3)

cat("\nMortality parameters:\n")
print(round(out_CP_GM_final$coefficients[1:3, ], 4))

cat("\nConvergence (Rhat):\n")
print(out_CP_GM_final$convergence)

life_CP <- round(out_CP_GM_final$lifeTable$noCov$Mean, 4)
print(life_CP)
```

```{r plot-cp}
plot(out_CP_GM_final, col = "cornflowerblue", plot.type = "fancy")
```


# ============================================================
# WARM LOWLAND (ORFORD) ANALYSIS
# ============================================================

# 7. Prepare Orford data
```{r prep-or}
dataOR <- data %>% filter(site == "OR") %>% select(-site)
inputMat_OR <- prepare_basta_input(dataOR, study_start = 2000, study_end = 2024)
```

# 8. Model selection — compare mortality functions (OR)

Fit all combinations of mortality model x shape to identify the best-fitting
parametric survival model. This is a coarse screen using shorter MCMC chains.

```{r basta-or-multi, eval=FALSE}
set.seed(123)
multiout_OR <- multibasta(
  object = inputMat_OR, studyStart = 2000, studyEnd = 2024,
  models = c("EX", "GO", "WE", "LO"),
  shapes  = c("simple", "Makeham"),
  nsim = 4, niter = 10000, burnin = 2000, thinning = 10, ncpus = 4)

saveRDS(multiout_OR, here("basta_outputs", "multiout_OR.rds"))
```

```{r load-multi-or}
multiout_OR <- readRDS(here("basta_outputs", "multiout_OR.rds"))
summary(multiout_OR, digits = 3)
# Result: WE simple selected as best model by DIC (4852), then WE Makeham (4863)
```

# 9. Validate recapture assumption (OR)

BaSTA assumes constant recapture probability across years. Before running the
final model, we test this by comparing the DIC of the best mortality model
(Weibull simple) under constant vs. time-varying recapture. Lower DIC supports
the better-fitting recapture structure.

```{r recapture-dic-or, eval=FALSE}
out_OR_WE_const <- basta(
  object = inputMat_OR, studyStart = 2000, studyEnd = 2024,
  model = "WE", shape = "simple", minAge = 2,
  nsim = 4, niter = 50000, burnin = 5001, thinning = 20,
  updateJumps = TRUE)

out_OR_WE_tv <- basta(
  object = inputMat_OR, studyStart = 2000, studyEnd = 2024,
  model = "WE", shape = "simple", minAge = 2,
  recaptureType = "year",
  nsim = 4, niter = 50000, burnin = 5001, thinning = 20,
  updateJumps = TRUE)

saveRDS(out_OR_WE_const, here("basta_outputs", "out_OR_WE_const.rds"))
saveRDS(out_OR_WE_tv,    here("basta_outputs", "out_OR_WE_tv.rds"))
```

```{r recapture-dic-or-results}
out_OR_WE_const <- readRDS(here("basta_outputs", "out_OR_WE_const.rds"))
out_OR_WE_tv    <- readRDS(here("basta_outputs", "out_OR_WE_tv.rds"))

cat("OR — Constant recapture DIC:", out_OR_WE_const$DIC["DIC"], "\n")
cat("OR — Time-varying recapture DIC:", out_OR_WE_tv$DIC["DIC"], "\n")

cat("Constant recapture:\n")
print(out_OR_WE_const$coefficients[1:3, ])

cat("\nTime-varying recapture:\n")
print(out_OR_WE_tv$coefficients[1:3, ])

# Time-varying fits best
```

# 10. Final model — Weibull simple (OR) with time-varying recapture

Run with longer chains for robust posterior estimates and life table derivation.

```{r basta-or-final, eval=FALSE}
set.seed(123)
out_OR_WE_final <- basta(
  object = inputMat_OR, studyStart = 2000, studyEnd = 2024,
  model = "WE", shape = "simple", minAge = 2,
  recaptureType = "year",
  nsim = 4, niter = 50000, burnin = 5001, thinning = 20,
  updateJumps = TRUE, lifeTable = TRUE)

saveRDS(out_OR_WE_final, here("basta_outputs", "out_OR_WE_final.rds"))
```

```{r or-results}
out_OR_WE_final <- readRDS(here("basta_outputs", "out_OR_WE_final.rds"))
summary(out_OR_WE_final, digits = 3)

cat("\nMortality parameters:\n")
print(round(out_OR_WE_final$coefficients[1:3, ], 4))

cat("\nConvergence (Rhat):\n")
print(out_OR_WE_final$convergence)

life_OR <- round(out_OR_WE_final$lifeTable$noCov$Mean, 4)
print(life_OR)
```

```{r plot-or}
plot(out_OR_WE_final, col = "coral1", plot.type = "fancy")
```


# ============================================================
# CROSS-POPULATION COMPARISON
# ============================================================

# 11. Mortality hazard comparison figure
```{r mortality-comparison}
extract_mort_curve_full <- function(basta_out, site_name, max_age = 12) {
  ages <- basta_out$x
  idx <- ages <= max_age
  data.frame(
    age    = ages[idx],
    mean = basta_out$mort$nocov["Mean", idx],
    lower  = basta_out$mort$nocov["2.5%", idx],
    upper  = basta_out$mort$nocov["97.5%", idx],
    site   = site_name)
}

haz_data <- bind_rows(
  extract_mort_curve_full(out_OR_WE_final, "Warm lowland"),
  extract_mort_curve_full(out_CP_GM_final, "Cool highland"))

p_haz_overlay <- ggplot(haz_data, aes(x = age, y = mean, color = site, fill = site)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, color = NA) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = site_colors) +
  scale_fill_manual(values = site_colors) +
  scale_x_continuous(breaks = seq(2, 12, 2)) +
  labs(y = "Mortality hazard \u03bc(x)", x = "Age (years)") +
  theme_publication

p_haz_overlay

ggsave(here("outputs", "figures", "Figure_mortality_comparison.png"),
       p_haz_overlay, width = 6, height = 5, dpi = 600)
ggsave(here("outputs", "figures", "Figure_mortality_comparison.pdf"),
       p_haz_overlay, width = 6, height = 5)
```

# 12. Life expectancy and mortality at key ages
```{r extract-model-agnostic}
# 1. Life expectancy at maturity from life tables
life_OR_WE <- out_OR_WE_final$lifeTable$noCov$Mean
life_CP_GM <- out_CP_GM_final$lifeTable$noCov$Mean

le_OR <- life_OR_WE$ex[life_OR_WE$Ages == 2]
le_CP <- life_CP_GM$ex[life_CP_GM$Ages == 2]
cat("Life expectancy at age 2 - Warm lowland:", round(le_OR, 2), "\n")
cat("Life expectancy at age 2 - Cool highland:", round(le_CP, 2), "\n")

# 2. Mortality at specific ages from posterior
ages_of_interest <- c(2, 4, 6, 8, 10, 12)

extract_mort_at_ages <- function(basta_out, ages_of_interest) {
  all_ages <- basta_out$x
  idx <- match(ages_of_interest, all_ages)
  data.frame(
    age   = ages_of_interest,
    mean  = basta_out$mort$nocov["Mean", idx],
    lower = basta_out$mort$nocov["2.5%", idx],
    upper = basta_out$mort$nocov["97.5%", idx])
}

mort_OR_ages <- extract_mort_at_ages(out_OR_WE_final, ages_of_interest)
mort_CP_ages <- extract_mort_at_ages(out_CP_GM_final, ages_of_interest)

cat("\nWarm lowland mortality at key ages (Weibull simple):\n")
print(round(mort_OR_ages, 4))
cat("\nCool highland mortality at key ages (Gompertz-Makeham):\n")
print(round(mort_CP_ages, 4))

# 3. Fold-increase in mortality from age 2 to age 12
mort_OR_age2  <- mort_OR_ages$mean[mort_OR_ages$age == 2]
mort_OR_age12 <- mort_OR_ages$mean[mort_OR_ages$age == 12]
mort_CP_age2  <- mort_CP_ages$mean[mort_CP_ages$age == 2]
mort_CP_age12 <- mort_CP_ages$mean[mort_CP_ages$age == 12]

cat("\nFold-increase in mortality age 2 to 12:\n")
cat("Warm lowland:", round(mort_OR_age12 / mort_OR_age2, 2), "\n")
cat("Cool highland:", round(mort_CP_age12 / mort_CP_age2, 2), "\n")

# 4. Confirm senescence in OR: Weibull b1 > 1 indicates increasing hazard
cat("\nWeibull simple parameters (OR):\n")
print(round(out_OR_WE_final$coefficients[1:2,
      c("Mean", "Lower95%CI", "Upper95%CI")], 4))
# b1 > 1 = senescence; check CI excludes 1

# 5. Posterior overlap on mortality trajectories at key ages
overlap_check <- data.frame(
  age      = ages_of_interest,
  OR_mean  = mort_OR_ages$mean,
  OR_lower = mort_OR_ages$lower,
  OR_upper = mort_OR_ages$upper,
  CP_mean  = mort_CP_ages$mean,
  CP_lower = mort_CP_ages$lower,
  CP_upper = mort_CP_ages$upper)

overlap_check$CIs_overlap <-
  overlap_check$OR_lower < overlap_check$CP_upper &
  overlap_check$CP_lower < overlap_check$OR_upper

print(round(overlap_check, 4))
```

# 13. Median lifespan (age at which survival drops below 0.5)
```{r median-lifespan}
derive_median_lifespan <- function(basta_out, site_name) {
  ages     <- basta_out$x
  mort_mean <- basta_out$mort$nocov["Mean", ]
  
  # Cumulative survival from hazard (numerical integration)
  surv <- exp(-cumsum(mort_mean * c(ages[1], diff(ages))))
  
  # Age at which survival first drops below 0.5
  if (any(surv <= 0.5)) {
    median_ls <- ages[which(surv <= 0.5)[1]]
  } else {
    median_ls <- max(ages)
    warning(site_name, ": survival never drops below 0.5 within observed age range")
  }
  
  cat(site_name, "median lifespan:", round(median_ls, 2), "years\n")
  return(median_ls)
}

median_OR <- derive_median_lifespan(out_OR_WE_final, "Warm lowland")
median_CP <- derive_median_lifespan(out_CP_GM_final, "Cool highland")
```


# ============================================================
# TABLES FOR MANUSCRIPT
# ============================================================

# 14. Parameter tables — presented separately as models differ between populations
```{r table-parameters}
# OR: Weibull simple (b0, b1)
# CP: Gompertz-Makeham (b0, b1, c)
# These are not directly comparable so are presented as separate tables


params_OR <- as.data.frame(out_OR_WE_final$coefficients) %>%
  rownames_to_column("Parameter") %>%
  filter(Parameter %in% c("b0", "b1")) %>%
  select(Parameter, Mean, StdErr, Lower = "Lower95%CI", Upper = "Upper95%CI",
         Rhat = "PotScaleReduc") %>%
  mutate(across(where(is.numeric), ~round(.x, 3)))

params_CP <- as.data.frame(out_CP_GM_final$coefficients) %>%
  rownames_to_column("Parameter") %>%
  filter(Parameter %in% c("b0", "b1", "c")) %>%
  select(Parameter, Mean, StdErr, Lower = "Lower95%CI", Upper = "Upper95%CI",
         Rhat = "PotScaleReduc") %>%
  mutate(across(where(is.numeric), ~round(.x, 3)))

ft_OR <- flextable(params_OR) %>%
  set_header_labels(StdErr = "SE", Lower = "Lower 95% CI", 
                    Upper = "Upper 95% CI", Rhat = "R-hat") %>%
  set_caption("Warm lowland (Weibull simple) mortality parameters.") %>%
  theme_vanilla() %>%
  align(align = "center", part = "all") %>%
  align(j = 1, align = "left", part = "body") %>%
  autofit()

ft_CP <- flextable(params_CP) %>%
  set_header_labels(StdErr = "SE", Lower = "Lower 95% CI", 
                    Upper = "Upper 95% CI", Rhat = "R-hat") %>%
  set_caption("Cool highland (Gompertz-Makeham) mortality parameters.") %>%
  theme_vanilla() %>%
  align(align = "center", part = "all") %>%
  align(j = 1, align = "left", part = "body") %>%
  autofit()

ft_OR
ft_CP

save_as_docx(ft_OR, path = here("outputs", "tables", "Table_params_warm_lowland.docx"))
save_as_docx(ft_CP, path = here("outputs", "tables", "Table_params_cool_highland.docx"))
```

# 15. Life tables (Supplementary Tables S12–S13)
```{r table-lifetables}
for (info in list(
  list(tbl = life_OR, pop = "Warm lowland",  cap = "Table S12",
       file = here("outputs", "tables", "Table_S12_lifetable_warm_lowland.docx")),
  list(tbl = life_CP, pop = "Cool highland", cap = "Table S13",
       file = here("outputs", "tables", "Table_S13_lifetable_cool_highland.docx")))) {

  tbl <- as.data.frame(info$tbl)
  tbl$Population <- info$pop

  ft_lt <- flextable(tbl) %>%
    set_caption(paste0(info$cap, ". Life table for ", info$pop,
                       " population.")) %>%
    theme_vanilla() %>%
    align(align = "center", part = "all") %>%
    autofit()

  save_as_docx(ft_lt, path = info$file)
}
```


# ============================================================
# SUPPLEMENTARY MATERIAL
# ============================================================

# 16. MCMC convergence diagnostics (Figure S1)
```{r trace-plots}
extract_trace <- function(model, param) {
  traces <- model$runs
  do.call(rbind, lapply(1:4, function(i) {
    theta_cols <- colnames(traces[[i]]$theta)
    if (param %in% theta_cols) {
      values <- traces[[i]]$theta[, param]
    } else if (param %in% names(traces[[i]])) {
      values <- traces[[i]][[param]][, 1]
    } else {
      stop(paste("Parameter", param, "not found"))
    }
    data.frame(iteration = seq_along(values), value = values,
               chain = paste0("Chain ", i))
  }))
}

make_trace_plot <- function(model, param, title = NULL, subtitle = NULL) {
  ggplot(extract_trace(model, param),
         aes(x = iteration, y = value, color = chain)) +
    geom_line(alpha = 0.7, linewidth = 0.3) +
    labs(title = title, subtitle = subtitle, x = "Iteration", y = "Value") +
    theme_bw() +
    theme(legend.position = "none",
          plot.title    = element_text(size = 10),
          plot.subtitle = element_text(size = 9))
}

# OR: Weibull simple — trace plots for b0, b1
p1 <- make_trace_plot(out_OR_WE_final, "b0", expression(b[0]))
p2 <- make_trace_plot(out_OR_WE_final, "b1", subtitle = expression(b[1])) 

# CP: Gompertz-Makeham — trace plots for b0, b1, c
p3 <- make_trace_plot(out_CP_GM_final, "b0", expression(b[0]))
p4 <- make_trace_plot(out_CP_GM_final, "b1", subtitle = expression(b[1]))
p5 <- make_trace_plot(out_CP_GM_final, "c",  subtitle = "c (Makeham)") 

fig_S1 <- grid.arrange(p1, p2, ncol = 2, nrow = 1)
fig_S2 <- grid.arrange( p3, p4, p5, ncol = 3, nrow = 1)

ggsave(here("outputs", "figures", "Figure_S1_convergence_OR.png"),
       fig_S1, width = 12, height = 7, dpi = 300)
ggsave(here("outputs", "figures", "Figure_S1_convergence_OR.pdf"),
       fig_S1, width = 12, height = 7)
ggsave(here("outputs", "figures", "Figure_S2_convergence_CP.png"),
       fig_S2, width = 12, height = 7, dpi = 300)
ggsave(here("outputs", "figures", "Figure_S2_convergence_CP.pdf"),
       fig_S2, width = 12, height = 7)
```

# 17. OR model robustness: Weibull simple vs Gompertz-Makeham

Weibull simple was the best fit for OR (Section 8). We tested Gompertz-Makeham
to assess whether it could be used instead to allow direct parameter comparison
with CP. However, the two models produce divergent mortality trajectories over
biologically relevant ages (<=14 years), confirming that Gompertz-Makeham is a
poor fit for OR and that population-specific best-fitting models must be used.
Cross-population comparisons therefore rely on model-agnostic outputs
(life expectancy, mortality at key ages) rather than parameter estimates.

```{r or-model-comparison, eval=FALSE}
out_OR_GM <- basta(
  object = inputMat_OR, studyStart = 2000, studyEnd = 2024,
  model = "GO", shape = "Makeham", minAge = 2,
  recaptureType = "year",
  nsim = 4, niter = 50000, burnin = 5001, thinning = 20,
  updateJumps = TRUE, lifeTable = TRUE)

saveRDS(out_OR_GM, here("basta_outputs", "out_OR_GM.rds"))
```

```{r or-model-comparison-results}
out_OR_GM <- readRDS(here("basta_outputs", "out_OR_GM.rds"))
out_OR_WE <- readRDS(here("basta_outputs", "out_OR_WE_final.rds"))

extract_mort_curve <- function(basta_out, model_label, max_age = 14) {
  data.frame(
    age       = basta_out$x,
    mortality = basta_out$mort$nocov["Mean", ],
    lower     = basta_out$mort$nocov["2.5%", ],
    upper     = basta_out$mort$nocov["97.5%", ],
    model     = model_label) %>%
    filter(age <= max_age)
}

mort_compare <- bind_rows(
  extract_mort_curve(out_OR_GM, "Gompertz-Makeham"),
  extract_mort_curve(out_OR_WE, "Weibull simple"))

ggplot(mort_compare, aes(x = age, y = mortality, color = model, fill = model)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, color = NA) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Gompertz-Makeham" = "#E69F00", "Weibull simple" = "#56B4E9")) +
  scale_fill_manual(values  = c("Gompertz-Makeham" = "#E69F00", "Weibull simple" = "#56B4E9")) +
  labs(x = "Age (years)", y = "Mortality rate \u03bc(x)",
       title = "Warm lowland: Gompertz-Makeham vs Weibull simple (ages 0-14)") +
  theme_classic() +
  theme(legend.position = "top")

# Quantify divergence between models
mort_wide <- inner_join(
  extract_mort_curve(out_OR_GM, "GM") %>% select(age, mort_GM = mortality),
  extract_mort_curve(out_OR_WE, "WE") %>% select(age, mort_WE = mortality),
  by = "age")

cat("Correlation (ages 0-14):", round(cor(mort_wide$mort_GM, mort_wide$mort_WE), 4), "\n")
cat("Mean abs % difference:",
    round(mean(abs(mort_wide$mort_GM - mort_wide$mort_WE) /
               ((mort_wide$mort_GM + mort_wide$mort_WE) / 2)) * 100, 2), "%\n")

# Both models detect senescence (b1 > 0 for GM; b1 > 1 for Weibull)
cat("\nGompertz-Makeham b1 (OR):\n")
print(out_OR_GM$coefficients["b1", c("Mean", "Lower95%CI", "Upper95%CI")])
cat("Weibull simple b1 (OR):\n")
print(out_OR_WE$coefficients["b1", c("Mean", "Lower95%CI", "Upper95%CI")])
```

# 18. Proxy for dead — gap analysis
```{r gap-analysis}
# Use a separate object to avoid overwriting data used in BaSTA analysis
# Remove 2023-2024 as too recent to assess the "returned after 2 years" pattern
data_gaps <- data %>%
  filter(year <= 2022) %>%
  # Retain rows with complete data on columns needed for gap analysis only
  filter(!is.na(femid) & !is.na(year) & !is.na(site)) %>%
  arrange(femid, site, year)

# Calculate year gaps for each individual
gap_analysis <- data_gaps %>%
  group_by(femid, site) %>%
  arrange(year) %>%
  mutate(
    next_year  = lead(year),
    year_gap   = next_year - year,
    is_last_observation = is.na(next_year)) %>%
  ungroup()

# Identify females that missed 2+ years but then returned
# year_gap >= 3: seen in year X and year X+3 means years X+1 and X+2 were missed
returned_after_2yrs <- gap_analysis %>%
  filter(year_gap >= 3) %>%
  select(femid, site, year, next_year, year_gap)

# Summary by site
summary_stats <- gap_analysis %>%
  group_by(site) %>%
  summarise(
    total_females       = n_distinct(femid),
    total_observations  = n(),
    mean_gap            = mean(year_gap[!is_last_observation], na.rm = TRUE),
    median_gap          = median(year_gap[!is_last_observation], na.rm = TRUE),
    max_gap             = max(year_gap[!is_last_observation], na.rm = TRUE),
    n_gaps_2plus_missed = sum(year_gap >= 3, na.rm = TRUE),  # 2+ years missed
    n_gaps_1yr_missed   = sum(year_gap == 2, na.rm = TRUE),  # 1 year missed
    prop_consecutive    = sum(year_gap == 1, na.rm = TRUE) / sum(!is_last_observation))

print(summary_stats)

cat("\nNumber of instances where females returned after 2+ years missing:\n")
print(table(returned_after_2yrs$site))

cat("\nFemales that returned after missing 2+ years:\n")
print(returned_after_2yrs)

# Recapture probability
recapture_prob <- gap_analysis %>%
  filter(!is_last_observation) %>%
  group_by(site) %>%
  summarise(
    prob_recap_next_year    = sum(year_gap == 1) / n(),
    prob_recap_within_2yrs  = sum(year_gap <= 2) / n())

cat("\nRecapture probabilities:\n")
print(recapture_prob)

# Distribution of maximum gap before returning, per individual
max_gaps <- gap_analysis %>%
  filter(!is_last_observation) %>%
  group_by(femid, site) %>%
  summarise(max_gap_before_return = max(year_gap, na.rm = TRUE)) %>%
  ungroup()

gap_distribution <- max_gaps %>%
  group_by(site) %>%
  count(max_gap_before_return) %>%
  mutate(prop = n / sum(n))

cat("\nDistribution of maximum gaps before returning:\n")
print(gap_distribution)

# Visualization
p_gaps <- ggplot(gap_distribution, 
                 aes(x = max_gap_before_return, y = prop, fill = site)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = site_colors) +
  labs(x = "Years missed before returning",
       y = "Proportion of females",
       title = "Maximum gap length before individuals returned") +
  theme_publication +
  theme(legend.position = "top")

p_gaps

ggsave(here("outputs", "figures", "Figure_gap_analysis.png"),
       p_gaps, width = 6, height = 5, dpi = 300)
ggsave(here("outputs", "figures", "Figure_gap_analysis.pdf"),
       p_gaps, width = 6, height = 5)

# Percentages for supplementary text
gap_analysis %>%
  filter(!is_last_observation) %>%
  group_by(site) %>%
  summarise(
    total_observable_gaps = n(),
    gaps_2plus_yrs_missed = sum(year_gap >= 3),
    percent_2plus         = round((sum(year_gap >= 3) / n()) * 100, 1))

# Temporal pattern of returns
returned_after_2yrs %>%
  group_by(site) %>%
  summarise(
    earliest_year = min(year),
    latest_year   = max(year),
    median_year   = median(year))
```