---
title: "Reproductive_senescence_BaFTA"
author: "Luisa Fitzpatrick"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 6)

```

# Setup

## Load packages

```{r packages}
library(here)
library(tidyverse)
library(cowplot)
library(ggplot2)
library(gridExtra)
library(snowfall)
library(BaFTA)
library(flextable)
library(officer)
library(lme4)
library(lmerTest)
library(emmeans)
```

# Plot style

```{r}
theme_publication <- theme_classic() +
  theme(
    axis.text  = element_text(size = 11),
    axis.title = element_text(size = 12),
    legend.text  = element_text(size = 11),
    legend.title = element_text(size = 12)
  )
```

## Load and prepare data

```{r load-data}
# Load data
data <- read.csv(here("data", "processed", "reproductive_data.csv"), 
                 na = c("", "NA", ".", "?"))[, c('Year', 'FemID', 'Site', 'Age', 'Clutch')]

# Rename columns
data <- data %>%
  rename(
    year = Year,
    indID = FemID,
    site = Site,
    nOffspring = Clutch)

# Format variables
data$indID <- as.factor(data$indID)
data$site <- as.factor(data$site)
data$Age <- as.integer(data$Age)
data$nOffspring <- as.integer(data$nOffspring)

# Remove missing data
data <- subset(data,
              !is.na(indID) & indID != 0 &
              !is.na(Age) & Age != 0 &
              !is.na(nOffspring) & nOffspring != 0) 
# NOTE: Zero litter sizes are excluded because non-gravid status, stillbirths, and mid-pregnancy reabsorptions were not consistently recorded throughout the study. Zeros therefore cannot be distinguished from missing data, so this analysis models conditional fertility (litter size given reproduction occurred).

head(data)

cat("Final n:", nrow(data), "\n")
```

## Subset by population

```{r subset-data}
dataCP <- subset(data, site == "CP")
dataOR <- subset(data, site == "OR")

cat("CP sample size:", nrow(dataCP), "\n")
cat("OR sample size:", nrow(dataOR), "\n")
```

# Visualize raw data

```{r raw-data-plots}
# CP population
p_raw_CP <- ggplot(dataCP, aes(x = Age, y = nOffspring)) +
  geom_point(alpha = 0.3, color = "cornflowerblue") +
  stat_summary(fun = mean, geom = "line", color = "cornflowerblue", size = 1) +
  labs(title = "Cool Highlands (CP)", x = "Age (years)", y = "Litter size") +
  theme_minimal()

# OR population
p_raw_OR <- ggplot(dataOR, aes(x = Age, y = nOffspring)) +
  geom_point(alpha = 0.3, color = "coral") +
  stat_summary(fun = mean, geom = "line", color = "coral", size = 1) +
  labs(title = "Warm Lowlands (OR)", x = "Age (years)", y = "Litter size") +
  theme_minimal()

# Display
plot_grid(p_raw_CP, p_raw_OR, ncol = 2)
```

**Pattern:** Both populations show early increase. Orford plateaus around 5. CP continues increasing and maybe plateaus at 10.

---

# Model Fitting - CP Population

## CP: Quadratic model

```{r bafta-cp-quad, eval=FALSE}
# Run model (long computation)
set.seed(123)

out_quad_CP <- bafta(object = dataCP, model = "quadratic",
                     dataType = "indivSimple", nsim = 4, ncpus = 4,
                     niter = 50000, burnin = 10000, thinning = 50)

# Save output
dir.create(here("outputs", "bafta_outputs"), showWarnings = FALSE)
saveRDS(out_quad_CP, here("outputs", "bafta_outputs", "out_quad_CP.rds"))
```

```{r load-cp-quad}
# Load saved model
out_quad_CP <- readRDS(here("outputs", "bafta_outputs", "out_quad_CP.rds"))
out_quad_CP
```

**Results:**
- Convergence: All parameter chains converged
- DIC = 4128.85
- Predictive loss: Good (Fit = 1312, Penalty = 4465, Deviance = 5777)
- Smallest Neff = 67 Critically low (should be over 400) - parameter estimates unreliable

```{r plot-cp-quad}
# Diagnostic plots
plot(out_quad_CP)
plot(out_quad_CP, type = "fertility")
plot(out_quad_CP, type = "predictive")
```

## CP: Gamma model

```{r bafta-cp-gamma, eval=FALSE}
# Run model (long computation)
set.seed(123)

out_gamma_CP <- bafta(object = dataCP, model = "gamma",
                      dataType = "indivSimple", nsim = 4, ncpus = 4,
                      niter = 50000, burnin = 10000, thinning = 50)

# Save output
saveRDS(out_gamma_CP, here("outputs", "bafta_outputs", "out_gamma_CP.rds"))
```

```{r load-cp-gamma}
# Load saved model
out_gamma_CP <- readRDS(here("outputs", "bafta_outputs", "out_gamma_CP.rds"))
out_gamma_CP
```

**Results:**
- Convergence: All parameter chains converged 
- DIC = 4237.84
- Predictive loss: Good (Fit = 1540, Penalty = 4358, Deviance = 5899)
- Smallest Neff = 791 - OK

```{r plot-cp-gamma}
# Diagnostic plots
plot(out_gamma_CP)
plot(out_gamma_CP, type = "fertility") #not good
plot(out_gamma_CP, type = "predictive")
```

## CP: Colchero-Muller model

```{r bafta-cp-cm, eval=FALSE}
# Run model (long computation)
set.seed(123)

out_CM_CP <- bafta(object = dataCP, model = "ColcheroMuller",
                   dataType = "indivSimple", nsim = 4, ncpus = 4,
                   niter = 50000, burnin = 10000, thinning = 50)

# Save output
saveRDS(out_CM_CP, here("outputs", "bafta_outputs", "out_CM_CP.rds"))
```

```{r load-cp-cm}
# Load saved model
out_CM_CP <- readRDS(here("outputs", "bafta_outputs", "out_CM_CP.rds"))
out_CM_CP
```

**Results:**
- Convergence: All parameter chains converged 
- DIC = 4128.57
- Predictive loss: Good (Fit = 1307, Penalty = 4465, Deviance = 5772)
- Smallest Neff = 614 - OK 

```{r plot-cp-cm}
# Diagnostic plots
plot(out_CM_CP)
plot(out_CM_CP, type = "fertility")
plot(out_CM_CP, type = "predictive")
```

## CP Model comparison

**Best model for CP:** Colchero-Muller (DIC = 4128.57)
- Essentially the same DIC as quadratic (which had very low neff for b2)
- Visually, both quadratic and CM fit data well
- CM model preferred due to better convergence

---

# Model Fitting - OR Population

## OR: Quadratic model

```{r bafta-or-quad, eval=FALSE}
# Run model (long computation)
set.seed(123)

out_quad_OR <- bafta(object = dataOR, model = "quadratic",
                     dataType = "indivSimple", nsim = 4, ncpus = 4,
                     niter = 50000, burnin = 10000, thinning = 50)

# Save output
saveRDS(out_quad_OR, here("outputs", "bafta_outputs", "out_quad_OR.rds"))
```

```{r load-or-quad}
# Load saved model
out_quad_OR <- readRDS(here("outputs", "bafta_outputs", "out_quad_OR.rds"))
out_quad_OR
```

**Results:**
- Convergence: All parameter chains converged 
- DIC = 2224.25
- Predictive loss: Good (Fit = 526, Penalty = 1866, Deviance = 2392)
- Smallest Neff = 664 - Ok 

```{r plot-or-quad}
# Diagnostic plots
plot(out_quad_OR)
plot(out_quad_OR, type = "fertility")
plot(out_quad_OR, type = "predictive")
```

## OR: Gamma model

```{r bafta-or-gamma, eval=FALSE}
# Run model (long computation)
set.seed(123)

out_gamma_OR <- bafta(object = dataOR, model = "gamma",
                      dataType = "indivSimple", nsim = 4, ncpus = 4,
                      niter = 50000, burnin = 10000, thinning = 50)

# Save output
saveRDS(out_gamma_OR, here("outputs", "bafta_outputs", "out_gamma_OR.rds"))
```

```{r load-or-gamma}
# Load saved model
out_gamma_OR <- readRDS(here("outputs", "bafta_outputs", "out_gamma_OR.rds"))
out_gamma_OR
```

**Results:**
- Convergence: All parameter chains converged 
- DIC = 2270.91
- Predictive loss: Good (Fit = 591, Penalty = 1810, Deviance = 2410)
- Smallest Neff = 825 - Good 

```{r plot-or-gamma}
# Diagnostic plots
plot(out_gamma_OR)
plot(out_gamma_OR, type = "fertility") # not good
plot(out_gamma_OR, type = "predictive")
```

## OR: Colchero-Muller model

```{r bafta-or-cm, eval=FALSE}
# Run model (long computation)
set.seed(123)

out_CM_OR <- bafta(object = dataOR, model = "ColcheroMuller",
                   dataType = "indivSimple", nsim = 4, ncpus = 4,
                   niter = 50000, burnin = 10000, thinning = 50)

# Save output
saveRDS(out_CM_OR, here("outputs", "bafta_outputs", "out_CM_OR.rds"))
```

```{r load-or-cm}
# Load saved model
out_CM_OR <- readRDS(here("outputs", "bafta_outputs", "out_CM_OR.rds"))
out_CM_OR
```

**Results:**
- Convergence: All parameter chains converged 
- DIC = 2226.68
- Predictive loss: Good (Fit = 529, Penalty = 1864, Deviance = 2393)
- Smallest Neff = 549 - OK 

```{r plot-or-cm}
# Diagnostic plots
plot(out_CM_OR)
plot(out_CM_OR, type = "fertility")
plot(out_CM_OR, type = "predictive")
```

## OR Model comparison

**Best model for OR:** CM (DIC = 2226.68)
- Essentially the same as quadratic DIC is 2.43 away
- Visually quadratic is the best fit, but CM is essentially equal - just poor fit at ages with low data
- CM model preferred for direct comparison with CP

---

# Final Model Selection

**Decision:** Use **Colchero-Muller model for both populations** to allow direct comparison of parameters across populations.

BaFTA results support previous modeling:
- No real difference in litter size trajectories between sites
- No evidence of senescence, rather increase and plateau pattern


# Extract Fertility Predictions

```{r extract-fertility}
# CP population
fert_CP <- data.frame(
  Age = out_CM_CP$x,
  Mean = out_CM_CP$fert[, "Mean"],
  Lower = out_CM_CP$fert[, "Lower"],
  Upper = out_CM_CP$fert[, "Upper"],
  Population = "Cool Highlands (CP)"
)

# OR population
fert_OR <- data.frame(
  Age = out_CM_OR$x,
  Mean = out_CM_OR$fert[, "Mean"],
  Lower = out_CM_OR$fert[, "Lower"],
  Upper = out_CM_OR$fert[, "Upper"],
  Population = "Warm Lowlands (OR)"
)

# Combine
fert_combined <- rbind(fert_CP, fert_OR)
```

# Visualization

## Fertility trajectories only

```{r plot-fertility-only}
ggplot(fert_combined, aes(x = Age, y = Mean, color = Population, fill = Population)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = Lower, ymax = Upper), alpha = 0.2, color = NA) +
  scale_color_manual(values = c("Cool Highlands (CP)" = "cornflowerblue", 
                                "Warm Lowlands (OR)" = "coral")) +
  scale_fill_manual(values = c("Cool Highlands (CP)" = "cornflowerblue", 
                               "Warm Lowlands (OR)" = "coral")) +
  labs(x = "Age (years)", 
       y = "Expected litter size",
       color = "Population",
       fill = "Population") +
  theme_classic() +
  theme(legend.position = "top")

```

## Fertility trajectories with raw data

```{r plot-fertility-with-data}
ggplot() +
  # Raw data points (semi-transparent)
  geom_jitter(data = dataCP, aes(x = Age, y = nOffspring), 
             color = "cornflowerblue", alpha = 0.2, size = 1) +
  geom_jitter(data = dataOR, aes(x = Age, y = nOffspring), 
             color = "coral", alpha = 0.2, size = 1) +
  # Fitted curves with confidence intervals
  geom_ribbon(data = fert_combined, 
              aes(x = Age, ymin = Lower, ymax = Upper, fill = Population), 
              alpha = 0.3) +
  geom_line(data = fert_combined, 
            aes(x = Age, y = Mean, color = Population), 
            linewidth = 1.2) +
  scale_color_manual(values = c("Cool Highlands (CP)" = "cornflowerblue", 
                                "Warm Lowlands (OR)" = "coral")) +
  scale_fill_manual(values = c("Cool Highlands (CP)" = "cornflowerblue", 
                               "Warm Lowlands (OR)" = "coral")) +
  labs(x = "Age (years)", 
       y = "Litter size",
       color = "Population",
       fill = "Population") +
  theme_classic() +
  theme(legend.position = "top")
```

## Fertility trajectories with parameter densities

```{r publication-figure}
# ============== DEFINE COLOURS (single source of truth) ==============
site_colors <- c("Cool Highlands (CP)" = "cornflowerblue", 
                 "Warm Lowlands (OR)"  = "coral")

# ============== EXTRACT KEY PARAMETER POSTERIORS ==============
params_OR <- as.data.frame(out_CM_OR$theta) %>%
  mutate(site = "Warm Lowlands (OR)")
params_CP <- as.data.frame(out_CM_CP$theta) %>%
  mutate(site = "Cool Highlands (CP)")

params_combined <- bind_rows(params_OR, params_CP) %>%
  pivot_longer(cols = c(b0, b3),
               names_to = "parameter", 
               values_to = "value")

# ============== PANEL A: FERTILITY TRAJECTORIES ==============
p_fertility <- ggplot(fert_combined, aes(x = Age, y = Mean, 
                                          color = Population, fill = Population)) +
  geom_ribbon(aes(ymin = Lower, ymax = Upper), alpha = 0.2, color = NA) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = site_colors) +
  scale_fill_manual(values = site_colors) +
  scale_x_continuous(breaks = seq(2, 12, 2), limits = c(2, 12)) +
  labs(y = "Expected litter size", x = "Age (years)") +
  theme_publication +
  theme(legend.position = "none")

# ============== PANEL B: KEY PARAMETER DISTRIBUTIONS ==============
p_params <- ggplot(params_combined, aes(x = value, fill = site, color = site)) +
  geom_density(alpha = 0.4, linewidth = 1) +
  facet_wrap(~parameter, scales = "free", ncol = 1,
             labeller = labeller(parameter = c(
               "b0" = "b0 (baseline fertility)",
               "b3" = "b3 (curvature parameter)"))) +
  scale_fill_manual(values = site_colors) +
  scale_color_manual(values = site_colors) +
  labs(x = "Parameter value", y = "Posterior density") +
  theme_publication +
  theme(legend.position = "none",
        strip.background = element_blank(),
        strip.text = element_text(size = 10, hjust = 0))

# ============== COMBINE PANELS ==============
combined_figure <- plot_grid(
  p_fertility, p_params,
  ncol = 1,
  labels = c("A", "B"),
  label_size = 14)

combined_figure


```

---

# Extract and Export Coefficients

```{r extract-coefficients}
# Extract coefficient tables
coef_CP <- out_CM_CP$coefficients
coef_OR <- out_CM_OR$coefficients

# View them
print("CP coefficients:")
print(coef_CP)

print("OR coefficients:")
print(coef_OR)
```

## Create coefficient table for Word

```{r coefficient-table}
# Function to round to significant figures
signif_format <- function(x, digits = 3) {
  signif(x, digits)
}

# Create a combined dataframe with UNIQUE column names
coef_table <- data.frame(
  Parameter = rownames(coef_CP),
  CP_Mean = signif_format(coef_CP[, "Mean"]),
  CP_SD = signif_format(coef_CP[, "SD"]),
  CP_Lower = signif_format(coef_CP[, "Lower"]),
  CP_Upper = signif_format(coef_CP[, "Upper"]),
  OR_Mean = signif_format(coef_OR[, "Mean"]),
  OR_SD = signif_format(coef_OR[, "SD"]),
  OR_Lower = signif_format(coef_OR[, "Lower"]),
  OR_Upper = signif_format(coef_OR[, "Upper"])
)

# Create flextable
ft <- flextable(coef_table) %>%
  set_header_labels(
    Parameter = "Parameter",
    CP_Mean = "Mean",
    CP_SD = "SD",
    CP_Lower = "Lower 95% CI",
    CP_Upper = "Upper 95% CI",
    OR_Mean = "Mean",
    OR_SD = "SD",
    OR_Lower = "Lower 95% CI",
    OR_Upper = "Upper 95% CI"
  ) %>%
  add_header_row(
    values = c("", "Cool Highlands (CP)", "Warm Lowlands (OR)"),
    colwidths = c(1, 4, 4)
  ) %>%
  align(align = "center", part = "all") %>%
  align(j = 1, align = "left", part = "body") %>%
  bold(part = "header") %>%
  fontsize(size = 10, part = "all") %>%
  font(fontname = "Times New Roman", part = "all") %>%
  border_outer(border = fp_border(width = 2), part = "all") %>%
  border_inner_h(border = fp_border(width = 1), part = "all") %>%
  border_inner_v(border = fp_border(width = 1), part = "all") %>%
  add_footer_lines("Values rounded to 3 significant figures.") %>%
  autofit()

ft


print(coef_table)
```

---

# Model Comparison Table

```{r model-comparison-table}
# Create model comparison table
model_comparison <- data.frame(
  Population = rep(c("Cool Highlands (CP)", "Warm Lowlands (OR)"), each = 3),
  Model = rep(c("Quadratic", "Gamma", "Colchero-Muller"), 2),
  DIC = c(out_quad_CP$DIC["DIC"], out_gamma_CP$DIC["DIC"], out_CM_CP$DIC["DIC"],
        out_quad_OR$DIC["DIC"], out_gamma_OR$DIC["DIC"], out_CM_OR$DIC["DIC"]),
  Predictive_Loss = c(out_quad_CP$PredLoss[1, "Deviance"],
                      out_gamma_CP$PredLoss[1, "Deviance"],
                      out_CM_CP$PredLoss[1, "Deviance"],
                      out_quad_OR$PredLoss[1, "Deviance"],
                      out_gamma_OR$PredLoss[1, "Deviance"],
                      out_CM_OR$PredLoss[1, "Deviance"])
)
# Display
knitr::kable(model_comparison, caption = "Model comparison for both populations")

# Save as supplementary table
ft_models <- flextable(model_comparison) %>%
  set_header_labels(
    Population = "Population",
    Model = "Model",
    DIC = "DIC",
    Predictive_Loss = "Predictive Loss"
  ) %>%
  bold(part = "header") %>%
  add_footer_lines("CP quadratic DIC should be interpreted with caution due to poor convergence.") %>%
  autofit()


ft_models
```

# Lifetime sucess LM
```{r lifetime reproductive output linear model}
# Load data
data1 <- read.csv(here("data", "processed", "reproductive_data.csv"), 
                 na = c("", "NA", ".", "?"))[, c('Year', 'FemID', 'Site', 'Age', 'Clutch')]

# Rename columns
data1 <- data1 %>%
  rename(
    year = Year,
    indID = FemID,
    site = Site,
    nOffspring = Clutch)

# Format variables
data1$indID <- as.factor(data1$indID)
data1$site <- as.factor(data1$site)
data1$Age <- as.integer(data1$Age)
data1$nOffspring <- as.integer(data1$nOffspring)

# Remove missing data
data1 <- subset(data1,
              !is.na(indID) & indID != 0 &
              !is.na(Age) & Age != 0 &
              !is.na(nOffspring) & nOffspring != 0) 

# Calculate lifespan
data1 <- data1 %>% 
  group_by(indID) %>% 
  mutate(Lifespan = max(Age, na.rm = TRUE)) %>%
  ungroup()

# Assign cohort
data1$cohort <- (data1$year - data1$Age)
data1$cohort <- as.integer(data1$year - data1$Age)


# Get total reproduction AND cohort for each female
totrepro_full <- data1 %>%
  group_by(indID) %>%
  summarise(
    total_offspring = sum(nOffspring),
    cohort = first(cohort),  
    site = first(site))

# Fit linear mixed model
totalrepro <- lmer(total_offspring ~ site + (1|cohort), data = totrepro_full)

# View results
summary(totalrepro)

anova(totalrepro)

emmeans(totalrepro, ~ site)

```


