---
title: "Growth Curve Analysis 2025"
author: "Luisa Fitzpatrick"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
  html_document: default
  word_document: default
---

```{r load-packages, message = FALSE, warning = FALSE}
library(here)
library(nlme)
library(ggplot2)
library(knitr)
library(dplyr)
library(effsize)
library(flextable)
library(officer)
# Note: MASS is used via MASS::mvrnorm() in the bootstrap functions
```

## Prepare data

```{r prepare-data}
# Read full dataset including juveniles
growth_data_all <- read.csv(here("data", "processed", "growth_data.csv"), 
                        na.strings = c("", ",", "NA", "."), 
                        header = TRUE)

# Filter to adults (age >= 2) for von Bertalanffy models
# Age 2 is the approximate age of first reproduction in this species,
# so VB models are fitted to post-maturation growth only
growth_data <- growth_data_all %>% 
  filter(!is.na(FSVLAFTER), !is.na(SITE.), 
         !is.na(Age), !is.na(FEMID.),
         Age >= 2)

# Check sample sizes by age and site
table(growth_data$Age, growth_data$SITE.)

# Separate by site for site-specific model fitting
growth_data_OR <- growth_data %>% filter(SITE. == "OR")
growth_data_CP <- growth_data %>% filter(SITE. == "CP")
```

## Summary statistics

```{r summary-stats}
summary_stats <- tibble(
  Population = c("Overall", "Warm lowland (OR)", "Cool highland (CP)"),
  Observations = c(nrow(growth_data), nrow(growth_data_OR), nrow(growth_data_CP)),
  Individuals = c(length(unique(growth_data$FEMID.)),
                  length(unique(growth_data_OR$FEMID.)),
                  length(unique(growth_data_CP$FEMID.))),
  Age_min = c(min(growth_data$Age), min(growth_data_OR$Age), min(growth_data_CP$Age)),
  Age_max = c(max(growth_data$Age), max(growth_data_OR$Age), max(growth_data_CP$Age)),
  Mean_obs_per_ind = c(
    round(mean(table(growth_data$FEMID.)), 1),
    round(mean(table(growth_data_OR$FEMID.)), 1),
    round(mean(table(growth_data_CP$FEMID.)), 1)),
  Obs_range_per_ind = c(
    paste0(min(table(growth_data$FEMID.)), "-", max(table(growth_data$FEMID.))),
    paste0(min(table(growth_data_OR$FEMID.)), "-", max(table(growth_data_OR$FEMID.))),
    paste0(min(table(growth_data_CP$FEMID.)), "-", max(table(growth_data_CP$FEMID.)))))

kable(summary_stats, digits = 1, 
      caption = "Summary statistics for adult growth data (age >= 2)")
```

## Fit mixed-effects von Bertalanffy models

Von Bertalanffy growth models were fitted using nonlinear mixed effects (nlme), with
individual identity (FEMID) as a random effect on the asymptotic size parameter (Linf).
This accounts for repeated measurements on the same individual and allows individuals
to vary in their asymptotic size while sharing common growth rate (K) and theoretical
age at zero size (t0) parameters at the population level. Models fitted by maximum
likelihood to allow AIC comparison.

```{r fit-vb-OR}
# ORFORD (warm lowland population)
fit_vb_OR <- nlme(
  FSVLAFTER ~ Linf * (1 - exp(-K * (Age - t0))),
  data = growth_data_OR,
  fixed = Linf + K + t0 ~ 1,
  random = Linf ~ 1 | FEMID.,
  start = c(Linf = 74, K = 0.044, t0 = 0),
  control = nlmeControl(maxIter = 200, msMaxIter = 200))

summary(fit_vb_OR)
intervals(fit_vb_OR)  # 95% CIs for fixed effects
```

```{r fit-vb-CP}
# CENTRAL PLATEAU (cool highland population)
fit_vb_CP <- nlme(
  FSVLAFTER ~ Linf * (1 - exp(-K * (Age - t0))),
  data = growth_data_CP,
  fixed = Linf + K + t0 ~ 1,
  random = Linf ~ 1 | FEMID.,
  start = c(Linf = 89, K = 0.044, t0 = 0),
  control = nlmeControl(maxIter = 200, msMaxIter = 200))

summary(fit_vb_CP)
intervals(fit_vb_CP)
```

## Model diagnostics

```{r diagnostic-function}
# Function to produce standard diagnostic plots for nlme models
check_nlme_diagnostics <- function(model, title) {
  
  par(mfrow = c(2, 2))
  
  # 1. Residuals vs Fitted - checks for systematic patterns (non-linearity, bias)
  plot(fitted(model), residuals(model, type = "pearson"),
       xlab = "Fitted values", ylab = "Pearson residuals",
       main = paste(title, "- Residuals vs Fitted"))
  abline(h = 0, col = "red", lty = 2)
  lines(lowess(fitted(model), residuals(model, type = "pearson")), 
        col = "blue", lwd = 2)
  
  # 2. QQ plot - checks normality of residuals
  qqnorm(residuals(model, type = "pearson"), 
         main = paste(title, "- Normal Q-Q"))
  qqline(residuals(model, type = "pearson"), col = "red")
  
  # 3. Scale-Location - checks homoscedasticity (constant variance)
  plot(fitted(model), sqrt(abs(residuals(model, type = "pearson"))),
       xlab = "Fitted values", ylab = "sqrt|Pearson residuals|",
       main = paste(title, "- Scale-Location"))
  lines(lowess(fitted(model), sqrt(abs(residuals(model, type = "pearson")))), 
        col = "blue", lwd = 2)
  
  # 4. Histogram - visualises residual distribution
  hist(residuals(model, type = "pearson"), breaks = 20,
       main = paste(title, "- Residual Distribution"),
       xlab = "Pearson residuals")
  
  par(mfrow = c(1, 1))
  
  # Shapiro-Wilk test for normality of residuals
  # Note: at large sample sizes (n > 400) this test is highly sensitive and
  # significant results should be interpreted alongside visual diagnostics
  cat("\n", title, "- Shapiro-Wilk test for normality:\n")
  print(shapiro.test(residuals(model, type = "pearson")))
}
```

```{r diagnostics-OR}
check_nlme_diagnostics(fit_vb_OR, "OR")
# OR diagnostics: Shapiro-Wilk is significant (p = 0.001) but visual inspection
# of the QQ plot shows only minor deviation at the tails, with residuals 
# following the theoretical line closely across the bulk of the distribution.
# Given the large sample size (n = 467), this level of Shapiro-Wilk sensitivity
# is expected and does not indicate a meaningful violation. Residuals vs fitted
# and scale-location plots show no systematic patterns.
```

```{r diagnostics-CP}
check_nlme_diagnostics(fit_vb_CP, "CP")
# CP diagnostics: Both Shapiro-Wilk (p = 2.9e-13) and visual inspection indicate
# more substantial departure from normality, with heavy tails on the QQ plot and
# a possible outlier at ~4 Pearson residuals. The scale-location plot suggests
# mild heteroscedasticity (increasing variance with fitted values).
# Testing variance structures below to assess whether this can be improved.
```

```{r diagnostics-random-effects}
# Check normality of random effects (individual-level Linf deviations)
# These should also be approximately normal for valid inference
par(mfrow = c(1, 2))
qqnorm(ranef(fit_vb_OR)[,1], main = "OR - Random Effects Q-Q")
qqline(ranef(fit_vb_OR)[,1], col = "red")
qqnorm(ranef(fit_vb_CP)[,1], main = "CP - Random Effects Q-Q")
qqline(ranef(fit_vb_CP)[,1], col = "red")
par(mfrow = c(1, 1))
# Random effects are approximately normally distributed for both populations.
```

## Test variance structures for CP heteroscedasticity

The CP model showed mild heteroscedasticity in residual diagnostics.
Two variance structures were tested to assess whether explicit variance
modelling could improve fit: varPower and varExp (both as functions of
fitted values).
```{r CP-variance-structures}
# Model 2: variance as power function of fitted values
fit_vb_CP_power <- nlme(
  FSVLAFTER ~ Linf * (1 - exp(-K * (Age - t0))),
  data = growth_data_CP,
  fixed = Linf + K + t0 ~ 1,
  random = Linf ~ 1 | FEMID.,
  weights = varPower(form = ~fitted(.)),
  start = c(Linf = 89, K = 0.044, t0 = 0),
  control = nlmeControl(maxIter = 200, msMaxIter = 200))
# Note: model produced a convergence warning during fitting

# Model 3 (varExp) failed to converge entirely (maxIter = 200 reached),
# indicating the exponential variance structure is not supported by these data.

# Compare base model vs power variance model
AIC(fit_vb_CP, fit_vb_CP_power)
summary(fit_vb_CP_power)

# Assessment: despite deltaAIC ~ 5 in favour of varPower, the model produced
# a degenerate solution - residual SD collapsed to ~0 (1.59e-07) with an
# implausibly large power parameter (~3.78), indicating the variance function
# absorbed all residual variance rather than capturing meaningful heteroscedasticity.
# The convergence warning and complete failure of varExp to converge further
# indicate this heteroscedasticity is not sufficiently structured to support
# explicit variance modelling.
# Base model retained; mild heteroscedasticity noted as a limitation.
# Fixed effect estimates are unlikely to be materially biased (n = 568).
fit_vb_CP_final <- fit_vb_CP
```


## Model selection: compare VB, logistic, and Gompertz

Three commonly used growth functions were compared by AIC to confirm VB as the
most appropriate functional form for each population.

```{r model-comparison}
# Logistic growth: FSVLAFTER = Linf / (1 + exp(-K * (Age - t0)))
fit_log_OR <- nlme(
  FSVLAFTER ~ Linf / (1 + exp(-K * (Age - t0))),
  data = growth_data_OR,
  fixed = Linf + K + t0 ~ 1,
  random = Linf ~ 1 | FEMID.,
  start = c(Linf = 74, K = 0.5, t0 = 1),
  control = nlmeControl(maxIter = 200, msMaxIter = 200))

fit_log_CP <- nlme(
  FSVLAFTER ~ Linf / (1 + exp(-K * (Age - t0))),
  data = growth_data_CP,
  fixed = Linf + K + t0 ~ 1,
  random = Linf ~ 1 | FEMID.,
  start = c(Linf = 89, K = 0.5, t0 = 1),
  control = nlmeControl(maxIter = 200, msMaxIter = 200))

# Gompertz growth: FSVLAFTER = Linf * exp(-exp(-K * (Age - t0)))
fit_gomp_OR <- nlme(
  FSVLAFTER ~ Linf * exp(-exp(-K * (Age - t0))),
  data = growth_data_OR,
  fixed = Linf + K + t0 ~ 1,
  random = Linf ~ 1 | FEMID.,
  start = c(Linf = 74, K = 0.5, t0 = 0),
  control = nlmeControl(maxIter = 200, msMaxIter = 200))

fit_gomp_CP <- nlme(
  FSVLAFTER ~ Linf * exp(-exp(-K * (Age - t0))),
  data = growth_data_CP,
  fixed = Linf + K + t0 ~ 1,
  random = Linf ~ 1 | FEMID.,
  start = c(Linf = 89, K = 0.5, t0 = 0),
  control = nlmeControl(maxIter = 200, msMaxIter = 200))

# AIC comparison within each population
AIC(fit_vb_OR, fit_log_OR, fit_gomp_OR)
AIC(fit_vb_CP, fit_log_CP, fit_gomp_CP)

# All three models have deltaAIC < 2 within each population, indicating
# they are statistically indistinguishable in fit. VB was selected as the
# preferred model on the basis of biological convention (widely used for
# reptile growth) and interpretability of parameters (Linf, K, t0),
# not statistical superiority.
```

## Site comparison: combined model

A single combined model with site-specific fixed effects for all three parameters
was used to formally test whether growth parameters differed between populations.
CP is the reference level (alphabetical ordering), so intercepts represent CP
estimates and SiteOR coefficients represent OR deviations from CP.

```{r combined-model}
growth_data_both <- growth_data %>%
  mutate(Site = factor(SITE.))
# Note: factor() will set CP as reference level (alphabetical order)
# Intercepts = CP estimates; SiteOR coefficients = difference (OR - CP)

fit_combined <- nlme(
  FSVLAFTER ~ Linf * (1 - exp(-K * (Age - t0))),
  data = growth_data_both,
  fixed = list(Linf ~ Site, K ~ Site, t0 ~ Site),
  random = Linf ~ 1 | FEMID.,
  start = c(81, -12, 0.41, -0.13, -2.35, -2.83),
  # start values: Linf_CP, Linf diff(OR-CP), K_CP, K diff(OR-CP), t0_CP, t0 diff(OR-CP)
  control = nlmeControl(maxIter = 200, msMaxIter = 200))

summary(fit_combined)

# Sequential F-tests for significance of site differences in each parameter
anova(fit_combined)
# CP has significantly larger asymptotic size (Linf) and faster growth rate (K) than OR.
# t0 also differs significantly between sites, though the effect is smaller.
# OR approaches its (lower) asymptote more slowly than CP approaches its (larger) asymptote.
```

## Results table

```{r results-table}

# --- Table 1: VB parameters with 95% CIs ---
results_table <- tibble(
  Population = rep(c("Warm lowland (OR)", "Cool highland (CP)"), each = 3),
  Parameter = rep(c("Linf (mm)", "K (year-1)", "t0 (year)"), 2),
  Estimate = c(fixef(fit_vb_OR), fixef(fit_vb_CP_final)),
  Lower_95CI = c(intervals(fit_vb_OR)$fixed[,"lower"], 
                 intervals(fit_vb_CP_final)$fixed[,"lower"]),
  Upper_95CI = c(intervals(fit_vb_OR)$fixed[,"upper"], 
                 intervals(fit_vb_CP_final)$fixed[,"upper"])) %>%
  mutate(across(c(Estimate, Lower_95CI, Upper_95CI), ~round(., 3)))

# --- Table 2: ANOVA results for site differences ---
anova_results <- as.data.frame(anova(fit_combined)) %>%
  tibble::rownames_to_column("Term") %>%
  filter(grepl("Site", Term)) %>%
  mutate(
    Parameter = c("Linf (mm)", "K (year-1)", "t0 (year)"),
    F_value = round(`F-value`, 1),
    p_value = c("<0.001", "<0.001", "0.017")
  ) %>%
  dplyr::select(Parameter, numDF, denDF, F_value, p_value)

```


## Early growth comparison (ages 0-2 years)

To determine whether population differences in growth are present from birth and
through the pre-reproductive phase, SVL was compared between populations at age
0.01 years (all offspring born in the lab are measured at birth, n = 1660 OR and 2250 CP)
and a subset of juveniles that were recaptured for survival analysis in Spring following 
their first winter dormant period (approximate age 0.75, n = 187).

Only these two discrete age classes are available prior to adulthood reflecting the measurement schedule of the long-term study. A linear mixed effects model would be preferable to account for individual identity, but of the offspring born (age 0.01, only approx 20% survive to age 2, so we cannot include ID). A linear model fit to two data points is unsuitable therefore we use a simple t-test between the two populations at the two juvenile age classes. As the sample size at birth is very large (n > 4000 across age classes), even small differences will be statistically
significant, and effect sizes should be interpreted alongside the model output.

```{r early-growth-data}

# --- Split into two age classes ---

birth_data <- growth_data_all %>% 
  filter(Age == 0.01) %>%
  mutate(Site = factor(SITE., levels = c("OR", "CP")))

juvenile_data <- growth_data_all %>% 
  filter(Age == 0.75) %>%
  mutate(Site = factor(SITE., levels = c("OR", "CP")))

# --- Check sample sizes ---
cat("Birth sample sizes:\n")
table(birth_data$Site)
cat("\nJuvenile sample sizes:\n")
table(juvenile_data$Site)

# --- Descriptive statistics ---
birth_data %>%
  group_by(Site) %>%
  summarise(n = n(),
            mean_SVL = mean(FSVLAFTER, na.rm = TRUE),
            sd_SVL = sd(FSVLAFTER, na.rm = TRUE),
            se_SVL = sd_SVL / sqrt(n))

juvenile_data %>%
  group_by(Site) %>%
  summarise(n = n(),
            mean_SVL = mean(FSVLAFTER, na.rm = TRUE),
            sd_SVL = sd(FSVLAFTER, na.rm = TRUE),
            se_SVL = sd_SVL / sqrt(n))

# --- Welch's t-tests ---
t_birth <- t.test(FSVLAFTER ~ Site, data = birth_data, var.equal = FALSE)
t_juvenile <- t.test(FSVLAFTER ~ Site, data = juvenile_data, var.equal = FALSE)

print(t_birth)
print(t_juvenile)

# --- Effect sizes (Cohen's d) ---

d_birth <- cohen.d(FSVLAFTER ~ Site, data = birth_data)
d_juvenile <- cohen.d(FSVLAFTER ~ Site, data = juvenile_data)

print(d_birth)
print(d_juvenile)

# --- Summary table ---
results_early_growth <- data.frame(
  Age_class = c("Birth (~0.01 years)", "Post-winter (~0.75 years)"),
  n_OR = c(nrow(filter(birth_data, Site == "OR")),
            nrow(filter(juvenile_data, Site == "OR"))),
  n_CP = c(nrow(filter(birth_data, Site == "CP")),
            nrow(filter(juvenile_data, Site == "CP"))),
  mean_OR = c(mean(filter(birth_data, Site == "OR")$FSVLAFTER),
               mean(filter(juvenile_data, Site == "OR")$FSVLAFTER)),
  mean_CP = c(mean(filter(birth_data, Site == "CP")$FSVLAFTER),
               mean(filter(juvenile_data, Site == "CP")$FSVLAFTER)),
  t_stat = c(t_birth$statistic, t_juvenile$statistic),
  df = c(t_birth$parameter, t_juvenile$parameter),
  p_value = c(t_birth$p.value, t_juvenile$p.value),
  cohens_d = c(d_birth$estimate, d_juvenile$estimate)
) %>%
  mutate(across(where(is.numeric), ~round(., 3)))

print(results_early_growth)

```


## Session information

```{r session-info}
sessionInfo()
```
