---
title: "Actuarial Senescence Analysis"
author: "Luisa Fitzpatrick"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(BaSTA)
library(cowplot)
library(ggplot2)
library(gridExtra)
library(lme4)
library(flextable)
library(officer)

# Shared plotting aesthetics
site_colors <- c("Warm lowland" = "coral1", "Cool highland" = "cornflowerblue")

theme_publication <- theme_classic(base_size = 12) +
  theme(
    axis.text = element_text(color = "black"),
    axis.line = element_line(color = "black"),
    legend.position = "none",
    plot.margin = margin(10, 10, 10, 10))
```

# 1. Load and prepare data
```{r load-data}
#setwd("C:/Users/luisaf/OneDrive - University of Tasmania/Postdoc/senescence paper/2023 senescence")

data <- read.csv("OcellatusNatPop_FullDataset_2025.csv",
                 na = c("", "NA", ".", "?")) %>%
  select(Year, FemID, Site, Age, DeathYear) %>%
  rename(year = Year, femid = FemID, site = Site, death = DeathYear) %>%
  mutate(
    birth = year - Age,
    femid = as.factor(femid),
    site  = as.factor(site),
    death = as.integer(death),
    birth = as.integer(birth)) %>%
  filter(!is.na(femid))
```

# 2. Helper functions
```{r helper-functions}
# Prepare BaSTA input matrix from site-level data
prepare_basta_input <- function(site_data, study_start, study_end) {
  site_data$birth[is.na(site_data$birth)] <- 0
  site_data$death[is.na(site_data$death)] <- 0
  
  Y <- CensusToCaptHist(ID = site_data$femid, d = site_data$year, dformat = "yyyy")
  
  birthDeath <- site_data %>%
    select(femid, birth, death) %>%
    distinct(femid, .keep_all = TRUE) %>%
    arrange(as.character(femid))
  
  n_year_cols <- ncol(Y) - 1  # minus the ID column
  inputMat <- as.data.frame(cbind(Y, birthDeath[, -1]))
  inputMat <- inputMat[, c(1, n_year_cols + 2, n_year_cols + 3, 2:(n_year_cols + 1))]
  
  cat("Total individuals:", nrow(inputMat), "\n")
  cat("Total capture events:", sum(Y[, -1]), "\n")
  
  DataCheck(inputMat, studyStart = study_start, studyEnd = study_end, silent = FALSE)
  
  return(inputMat)
}

  # Derive median lifespan (age at which survival drops below 0.5)
  lifespan_50 <- sapply(1:n_samples, function(j) {
    surv <- exp(-cumsum(haz_matrix[j, ]))
    if (any(surv <= 0.5)) ages[which(surv <= 0.5)[1]] else max(ages)
  })
  
  list(
    hazard = data.frame(
      age    = ages,
      median = apply(haz_matrix, 2, median),
      lower  = apply(haz_matrix, 2, quantile, 0.025),
      upper  = apply(haz_matrix, 2, quantile, 0.975),
      site   = site_name),
    lifespan = data.frame(lifespan_50 = lifespan_50, site = site_name)
  )

```


# ============================================================
# COOL HIGHLAND (CENTRAL PLATEAU) ANALYSIS
# ============================================================

# 3. Prepare Plateau data
```{r prep-cp}
dataCP <- data %>% filter(site == "CP") %>% select(-site)
inputMat_CP <- prepare_basta_input(dataCP, study_start = 2000, study_end = 2024)
```

# 4. Model selection — compare mortality functions (CP)
```{r basta-cp-multi, eval=FALSE}
set.seed(123)
multiout_CP <- multibasta(
  object = inputMat_CP, studyStart = 2000, studyEnd = 2024,
  models = c("EX", "GO", "WE", "LO"),
  shapes  = c("simple", "Makeham"),
  nsim = 4, niter = 10000, burnin = 2001, thinning = 10, ncpus = 4)

saveRDS(multiout_CP, "basta outputs/multiout_CP.rds")
```

```{r load-multi-cp}
multiout_CP <- readRDS("basta outputs/multiout_CP.rds")
summary(multiout_CP, digits = 3)
# Result: Gompertz-Makeham selected as best model by DIC
```

# 5. Validate recapture assumption (CP)
```{r recapture-dic-cp, eval=FALSE}
out_CP_const <- basta(
  object = inputMat_CP, studyStart = 2000, studyEnd = 2024,
  model = "GO", shape = "Makeham", minAge = 2,
  nsim = 4, niter = 50000, burnin = 5001, thinning = 20,
  updateJumps = TRUE, lifeTable = TRUE)

out_CP_tv <- basta(
  object = inputMat_CP, studyStart = 2000, studyEnd = 2024,
  model = "GO", shape = "Makeham", minAge = 2,
  recaptureType = "year",
  nsim = 4, niter = 50000, burnin = 5001, thinning = 20,
  updateJumps = TRUE)

saveRDS(out_CP_const, "basta outputs/out_CP_const.rds")
saveRDS(out_CP_tv,    "basta outputs/out_CP_tv.rds")
```

```{r recapture-dic-cp-results}
out_CP_const <- readRDS("basta outputs/out_CP_const.rds")
out_CP_tv <- readRDS("basta outputs/out_CP_tv.rds")

cat("CP — Constant recapture DIC:", out_CP_const$DIC["DIC"])
cat("CP — Time-varying recapture DIC:", out_CP_tv$DIC["DIC"])

cat("Constant recapture:\n")
print(out_CP_const$coefficients[1:3, ])

cat("\nTime-varying recapture:\n")
print(out_CP_tv$coefficients[1:3, ])

#time-varying fits best
```

# 6. Final model — Gompertz-Makeham (CP) with time-varying recapture
```{r basta-cp-final, eval=FALSE}
set.seed(123)
out_CP_GM_final <- basta(
  object = inputMat_CP, studyStart = 2000, studyEnd = 2024,
  model = "GO", shape = "Makeham", minAge = 2,
  recaptureType = "year",
  nsim = 4, niter = 50000, burnin = 5001, thinning = 20,
  updateJumps = TRUE, lifeTable = TRUE)

saveRDS(out_CP_GM_final, "basta outputs/out_CP_GM_final.rds")
```

```{r cp-results}
out_CP <- readRDS("basta outputs/out_CP.rds")
summary(out_CP_GM_final, digits = 3)

cat("\nMortality parameters:\n")
print(round(out_CP_GM_final$coefficients[1:3, ], 4))

cat("\nConvergence (Rhat):\n")
print(out_CP_GM_final$convergence)

life_CP <- round(out_CP_GM_final$lifeTable$noCov$Mean, 4)
print(life_CP)
```

```{r plot-cp}
plot(out_CP_GM_final, col = "cornflowerblue", plot.type = "fancy")
```

# ============================================================
# WARM LOWLAND (ORFORD) ANALYSIS
# ============================================================

# 7. Prepare Orford data
```{r prep-or}
dataOR <- data %>% filter(site == "OR") %>% select(-site)
inputMat <- prepare_basta_input(dataOR, study_start = 2000, study_end = 2024)
```

# 8. Model selection — compare mortality functions (OR)

Fit all combinations of mortality model × shape to identify the best-fitting
parametric survival model. This is a coarse screen using shorter MCMC chains.

```{r basta-or-multi, eval=FALSE}
set.seed(123)
multiout_OR <- multibasta(
  object = inputMat, studyStart = 2000, studyEnd = 2024,
  models = c("EX", "GO", "WE", "LO"),
  shapes  = c("simple", "Makeham"),
  nsim = 4, niter = 10000, burnin = 2001, thinning = 10, ncpus = 4)

saveRDS(multiout_OR, "basta outputs/multiout_OR.rds")
```

```{r load-multi-or}
multiout_OR <- readRDS("basta outputs/multiout_OR.rds")
summary(multiout_OR, digits = 3)
# Result: WE simple selected as best model by DIC (4852), then WE Makeham (4863)
# Note: We tested GO Makeham overlap with WE simple to try and improve direct comparability with CP, but 
# See Section 17 for curve comparison
```

# 9. Validate recapture assumption (OR)

BaSTA assumes constant recapture probability across years. Before running the
final model, we test this by comparing the DIC of the best mortality model
(Gompertz-Makeham) under constant vs. time-varying recapture. If constant
recapture yields a lower or similar DIC, the assumption is justified.

```{r recapture-dic-or, eval=FALSE}
# Constant recapture (default) — same as the final model run below
out_OR_WE_const <- basta(
  object = inputMat, studyStart = 2000, studyEnd = 2024,
  model = "WE", shape = "simple", minAge = 2,
  nsim = 4, niter = 50000, burnin = 5001, thinning = 20,
  updateJumps = TRUE, lifeTable = TRUE)

# Time-varying recapture — recapture probability estimated separately per year
out_OR_WE_tv <- basta(
  object = inputMat, studyStart = 2000, studyEnd = 2024,
  model = "WE", shape = "simple", minAge = 2,
  recaptureType = "year",
  nsim = 4, niter = 50000, burnin = 5001, thinning = 20,
  updateJumps = TRUE)

saveRDS(out_OR_WE_const, "basta outputs/out_OR_WE_const.rds")
saveRDS(out_OR_WE_tv,    "basta outputs/out_OR_WE_tv.rds")
```

```{r recapture-dic-or-results}
out_OR_WE_const    <- readRDS("basta outputs/out_OR_WE_const.rds")
out_OR_WE_tv <- readRDS("basta outputs/out_OR_WE_tv.rds")

cat("OR — Constant recapture DIC:", out_OR_WE_const$DIC["DIC"])
cat("OR — Time-varying recapture DIC:", out_OR_WE_tv$DIC["DIC"])

cat("Constant recapture:\n")
print(out_OR_WE_const$coefficients[1:3, ])

cat("\nTime-varying recapture:\n")
print(out_OR_WE_tv$coefficients[1:3, ])

# time-varying fits best
```

# 10. Final model — Weibull-Simple (OR) time-varying recapture

Run with longer chains for robust posterior estimates and life table derivation.

```{r basta-or-final, eval=FALSE}
set.seed(123)
out_OR_WE_final <- basta(
  object = inputMat, studyStart = 2000, studyEnd = 2024,
  model = "WE", shape = "simple", minAge = 2,
  recaptureType = "year",
  nsim = 4, niter = 50000, burnin = 5001, thinning = 20,
  updateJumps = TRUE, lifeTable = TRUE)

saveRDS(out_OR_WE_final, "basta outputs/out_OR_WE_final.rds")
```

```{r or-results}
out_OR_WE_final <- readRDS("basta outputs/out_OR_WE_final.rds")
summary(out_OR_WE_final, digits = 3)

# Parameter summary
cat("\nMortality parameters:\n")
print(round(out_OR_WE_final$coefficients[1:3, ], 4))

cat("\nConvergence (Rhat):\n")
print(out_OR_WE_final$convergence)

# Life table
life_OR <- round(out_OR_WE_final$lifeTable$noCov$Mean, 4)
print(life_OR)
```

```{r plot-or}
plot(out_OR_WE_final, col = "coral1", plot.type = "fancy")
```


# ============================================================
# CROSS-POPULATION COMPARISON
# ============================================================

# 11. Mortality hazard comparison figure
```{r mortality-comparison}

extract_mort_curve_full <- function(basta_out, site_name, max_age = 12) {
  ages <- basta_out$x
  idx <- ages <= max_age
  data.frame(
    age    = ages[idx],
    median = basta_out$mort$nocov["Mean", idx],
    lower  = basta_out$mort$nocov["2.5%", idx],
    upper  = basta_out$mort$nocov["97.5%", idx],
    site   = site_name)
}

haz_data <- bind_rows(
  extract_mort_curve_full(out_OR_WE_final, "Warm lowland"),
  extract_mort_curve_full(out_CP_GM_final, "Cool highland"))

p_haz_overlay <- ggplot(haz_data, aes(x = age, y = median, color = site, fill = site)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, color = NA) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = site_colors) +
  scale_fill_manual(values = site_colors) +
  scale_x_continuous(breaks = seq(2, 12, 2)) +
  labs(y = "Mortality hazard μ(x)", x = "Age (years)") +
  theme_publication

p_haz_overlay

ggsave("Figure_mortality_comparison.png", p_haz_overlay,
       width = 6, height = 5, dpi = 600)
ggsave("Figure_mortality_comparison.pdf", p_haz_overlay,
       width = 6, height = 5)
```



# 13. Median lifespan
```{r median-lifespan}
cat("Warm lowland median lifespan:", median(traj_OR$lifespan$lifespan_50), "years\n")
cat("Cool highland median lifespan:", median(traj_CP$lifespan$lifespan_50), "years\n")
```

#14. Population comparisons

As different models fit each population best we can't directly compare model parameters

```{r extract model-agnostic comparisons}
out_OR_WE_final <- readRDS("basta outputs/out_OR_WE_final.rds")
out_CP_GM_final <- readRDS("basta outputs/out_CP_GM_final.rds")

# 1. Life expectancy at maturity from life tables
life_OR_WE <- out_OR_WE_final$lifeTable$noCov$Mean
life_CP_GM <- out_CP_GM_final$lifeTable$noCov$Mean

# Find life expectancy at age 2 (ex column at age 2)
le_OR <- life_OR_WE$ex[life_OR_WE$Ages == 2]
le_CP <- life_CP_GM$ex[life_CP_GM$Ages == 2]
cat("Life expectancy at age 2 - Warm lowland:", round(le_OR, 2), "\n")
cat("Life expectancy at age 2 - Cool highland:", round(le_CP, 2), "\n")

# 2. Mortality at specific ages from posterior (model-agnostic trajectory comparison)
# For Weibull simple (OR): extract from mort object
ages_of_interest <- c(2, 4, 6, 8, 10, 12)

extract_mort_at_ages <- function(basta_out, ages_of_interest) {
  all_ages <- basta_out$x
  idx <- match(ages_of_interest, all_ages)
  data.frame(
    age   = ages_of_interest,
    mean  = basta_out$mort$nocov["Mean", idx],
    lower = basta_out$mort$nocov["2.5%", idx],
    upper = basta_out$mort$nocov["97.5%", idx])
}

mort_OR_ages <- extract_mort_at_ages(out_OR_WE_final, ages_of_interest)
mort_CP_ages <- extract_mort_at_ages(out_CP_GM_final, ages_of_interest)

cat("\nWarm lowland mortality at key ages (Weibull simple):\n")
print(round(mort_OR_ages, 4))
cat("\nCool highland mortality at key ages (Gompertz-Makeham):\n")
print(round(mort_CP_ages, 4))

# 3. Mortality rate at oldest observed age (age 12) - how much has it increased?
# This is a model-agnostic way to describe the magnitude of senescence
mort_OR_age2  <- mort_OR_ages$mean[mort_OR_ages$age == 2]
mort_OR_age12 <- mort_OR_ages$mean[mort_OR_ages$age == 12]
mort_CP_age2  <- mort_CP_ages$mean[mort_CP_ages$age == 2]
mort_CP_age12 <- mort_CP_ages$mean[mort_CP_ages$age == 12]

cat("\nFold-increase in mortality age 2 to 12:\n")
cat("Warm lowland:", round(mort_OR_age12 / mort_OR_age2, 2), "\n")
cat("Cool highland:", round(mort_CP_age12 / mort_CP_age2, 2), "\n")

# 4. Confirm senescence in OR using Weibull b1
# In BaSTA Weibull simple, b1 > 1 indicates increasing hazard (senescence)
cat("\nWeibull simple parameters (OR):\n")
print(round(out_OR_WE_final$coefficients[1:2, 
      c("Mean", "Lower95%CI", "Upper95%CI")], 4))
# b1 > 1 = senescence; check CI excludes 1

# 5. Posterior overlap on mortality trajectories 
# At each age, does OR CI overlap CP CI?
mort_OR_full <- extract_mort_at_ages(out_OR_WE_final, ages_of_interest)
mort_CP_full <- extract_mort_at_ages(out_CP_GM_final, ages_of_interest)

overlap_check <- data.frame(
  age = ages_of_interest,
  OR_mean = mort_OR_full$mean,
  OR_lower = mort_OR_full$lower,
  OR_upper = mort_OR_full$upper,
  CP_mean = mort_CP_full$mean,
  CP_lower = mort_CP_full$lower,
  CP_upper = mort_CP_full$upper)

overlap_check$CIs_overlap <- 
  overlap_check$OR_lower < overlap_check$CP_upper & 
  overlap_check$CP_lower < overlap_check$OR_upper

print(round(overlap_check, 4))
```


# ============================================================
# TABLES FOR MANUSCRIPT
# ============================================================

# 15. Parameter comparison table (Table 1)
```{r table-parameters}
# Summarise posteriors for each population
summarise_params <- function(basta_out) {
  p <- basta_out$params
  data.frame(
    Parameter = colnames(p),
    Mean      = apply(p, 2, mean),
    SD        = apply(p, 2, sd),
    Lower95CI = apply(p, 2, quantile, 0.025),
    Upper95CI = apply(p, 2, quantile, 0.975))
}

params_OR_summary <- summarise_params(out_OR_GM_final)
params_CP_summary <- summarise_params(out_CP_GM_final)

key_params <- c("c", "b0", "b1", "lambda")

table_for_word <- data.frame(
  Parameter = c("c", "b\u2080", "b\u2081", "\u03bb"),
  WL_Mean      = round(params_OR_summary$Mean[match(key_params, params_OR_summary$Parameter)], 4),
  WL_SD        = round(params_OR_summary$SD[match(key_params, params_OR_summary$Parameter)], 4),
  WL_Lower95CI = round(params_OR_summary$Lower95CI[match(key_params, params_OR_summary$Parameter)], 4),
  WL_Upper95CI = round(params_OR_summary$Upper95CI[match(key_params, params_OR_summary$Parameter)], 4),
  CH_Mean      = round(params_CP_summary$Mean[match(key_params, params_CP_summary$Parameter)], 4),
  CH_SD        = round(params_CP_summary$SD[match(key_params, params_CP_summary$Parameter)], 4),
  CH_Lower95CI = round(params_CP_summary$Lower95CI[match(key_params, params_CP_summary$Parameter)], 4),
  CH_Upper95CI = round(params_CP_summary$Upper95CI[match(key_params, params_CP_summary$Parameter)], 4))

ft <- flextable(table_for_word) %>%
  add_header_row(values = c("", "Warm Lowland", "Cool Highland"), colwidths = c(1, 4, 4)) %>%
  set_header_labels(
    Parameter = "Parameter",
    WL_Mean = "Mean", WL_SD = "SD", WL_Lower95CI = "Lower 95% CI", WL_Upper95CI = "Upper 95% CI",
    CH_Mean = "Mean", CH_SD = "SD", CH_Lower95CI = "Lower 95% CI", CH_Upper95CI = "Upper 95% CI") %>%
  set_caption("Table 1. Gompertz-Makeham parameter estimates for warm lowland and cool highland populations.") %>%
  theme_vanilla() %>%
  align(align = "center", part = "all") %>%
  autofit()

ft
save_as_docx(ft, path = "Table1_BaSTA_parameters.docx")
write.csv(table_for_word, "Table1_BaSTA_parameters.csv", row.names = FALSE)
```

# 16. Life tables (Supplementary Tables S12–S13)
```{r table-lifetables}
for (info in list(
  list(tbl = life_OR, pop = "Warm lowland",  cap = "Table S12", file = "Table_S12_lifetable_warm_lowland.docx"),
  list(tbl = life_CP, pop = "Cool highland", cap = "Table S13", file = "Table_S13_lifetable_cool_highland.docx"))) {
  
  tbl <- as.data.frame(info$tbl)
  tbl$Population <- info$pop
  
  ft_lt <- flextable(tbl) %>%
    set_caption(paste0(info$cap, ". Life table for ", info$pop,
                       " population (Gompertz-Makeham model).")) %>%
    theme_vanilla() %>%
    align(align = "center", part = "all") %>%
    autofit()
  
  save_as_docx(ft_lt, path = info$file)
}
```

# ============================================================
# SUPPLEMENTARY MATERIAL
# ============================================================

# 17. MCMC convergence diagnostics (Figure S1)
```{r trace-plots}
extract_trace <- function(model, param) {
  traces <- model$runs
  do.call(rbind, lapply(1:4, function(i) {
    theta_cols <- colnames(traces[[i]]$theta)
    if (param %in% theta_cols) {
      values <- traces[[i]]$theta[, param]
    } else if (param %in% names(traces[[i]])) {
      values <- traces[[i]][[param]][, 1]
    } else {
      stop(paste("Parameter", param, "not found"))
    }
    data.frame(iteration = seq_along(values), value = values,
               chain = paste0("Chain ", i))
  }))
}

# Build trace plots for both populations
make_trace_plot <- function(model, param, title = NULL, subtitle = NULL) {
  ggplot(extract_trace(model, param),
         aes(x = iteration, y = value, color = chain)) +
    geom_line(alpha = 0.7, linewidth = 0.3) +
    labs(title = title, subtitle = subtitle, x = "Iteration", y = "Value") +
    theme_bw() +
    theme(legend.position = "none",
          plot.title = element_text(size = 10),
          plot.subtitle = element_text(size = 9))
}

p1 <- make_trace_plot(out_OR_GM_final, "b0", "Warm lowland (Gompertz-Makeham)", expression(b[0]))
p2 <- make_trace_plot(out_OR_GM_final, "b1", subtitle = expression(b[1]))
p3 <- make_trace_plot(out_OR_GM_final, "c",  subtitle = "c (Makeham)") +
  theme(legend.position = "bottom")
p4 <- make_trace_plot(out_CP_GM_final, "b0", "Cool highland (Gompertz-Makeham)", expression(b[0]))
p5 <- make_trace_plot(out_CP_GM_final, "b1", subtitle = expression(b[1]))
p6 <- make_trace_plot(out_CP_GM_final, "c",  subtitle = "c (Makeham)") +
  theme(legend.position = "bottom")

fig_S1 <- grid.arrange(p1, p2, p3, p4, p5, p6, ncol = 3, nrow = 2)

ggsave("Figure_S1_convergence.png", fig_S1, width = 12, height = 7, dpi = 300)
ggsave("Figure_S1_convergence.pdf", fig_S1, width = 12, height = 7)
```


# 18. OR model robustness: Weibull simple vs Gompertz-Makeham

Weibull simple was the best fit in the initial model selection (Section 8), then Weibull Makeham. Gompertz makeham is the third best fit. We can not confirm that the two models (weibull simple and Gompertz Maleham) produce qualitatively similar mortality trajectories over biologically relevant ages (≤14 years), so will use different models for the two populations.

```{r or-model-comparison}
# Run Gompertz Makeham to compare 
out_OR_GM <- basta(
  object = inputMat, studyStart = 2000, studyEnd = 2024,
  model = "GO", shape = "Makeham", minAge = 2,
  recaptureType = "year",
  nsim = 4, niter = 50000, burnin = 5001, thinning = 20,
  updateJumps = TRUE, lifeTable = TRUE)

saveRDS(out_OR_GM, "basta outputs/out_OR_GM.rds")

out_OR_GM <- readRDS("basta outputs/out_OR_GM.rds")  

# Load or run Weibull simple for comparison
out_OR_WE <- basta(
  object = inputMat, studyStart = 2000, studyEnd = 2024,
  model = "WE", shape = "simple", minAge = 2,
  recaptureType = "year",
  nsim = 4, niter = 50000, burnin = 5001, thinning = 20,
  updateJumps = TRUE, lifeTable = TRUE)

saveRDS(out_OR_WE, "basta outputs/out_OR_WE.rds")

out_OR_WE <- readRDS("basta outputs/out_OR_WE.rds")


# Extract mortality curves from both models
extract_mort_curve <- function(basta_out, model_label, max_age = 14) {
  data.frame(
    age       = basta_out$x,
    mortality = basta_out$mort$nocov["Mean", ],
    lower     = basta_out$mort$nocov["2.5%", ],
    upper     = basta_out$mort$nocov["97.5%", ],
    model     = model_label) %>%
    filter(age <= max_age)
}

mort_compare <- bind_rows(
  extract_mort_curve(out_OR_GM, "Gompertz-Makeham"),
  extract_mort_curve(out_OR_WE,    "Weibull simple"))

ggplot(mort_compare, aes(x = age, y = mortality, color = model, fill = model)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, color = NA) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Gompertz-Makeham" = "#E69F00", "Weibull simple" = "#56B4E9")) +
  scale_fill_manual(values  = c("Gompertz-Makeham" = "#E69F00", "Weibull simple" = "#56B4E9")) +
  labs(x = "Age (years)", y = "Mortality rate μ(x)",
       title = "Warm lowland: Gompertz-Makeham vs Weibull simple (ages 0–14)") +
  theme_classic() +
  theme(legend.position = "top")

# Quantify similarity
mort_wide <- inner_join(
  extract_mort_curve(out_OR_GM, "GM") %>% select(age, mort_GM = mortality),
  extract_mort_curve(out_OR_WE, "WE")    %>% select(age, mort_WE = mortality),
  by = "age")

cat("Correlation (ages 0–14):", round(cor(mort_wide$mort_GM, mort_wide$mort_WE), 4), "\n")
cat("Mean abs % difference:",
    round(mean(abs(mort_wide$mort_GM - mort_wide$mort_WE) /
               ((mort_wide$mort_GM + mort_wide$mort_WE) / 2)) * 100, 2), "%\n")

# Both models detect senescence (b1 > 0)
cat("\nGompertz-Makeham b1:\n")
print(out_OR_GM$coefficients["b1", c("Mean", "Lower95%CI", "Upper95%CI")])
cat("Weibull b1:\n")
print(out_OR_WE$coefficients["b1", c("Mean", "Lower95%CI", "Upper95%CI")])
```

