---
title: "Growth Curve Analysis 2025"
author: "Luisa Fitzpatrick"
date: "r Sys.Date()"
output:
  word_document: default
  html_document: default
---

```{r load-packages, message = FALSE, warning = FALSE}
library(nlme)
library(ggplot2)
library(knitr)
library(dplyr)
# Note: MASS is used via MASS::mvrnorm() in the bootstrap functions
```

## Prepare data
```{r}
# Read and filter to ages 2+
growth_data <- read.csv("Growthcurvedata.csv", 
                        na.strings = c("", ",", "NA", "."), 
                        header = TRUE)

growth_data <- growth_data %>% 
  filter(!is.na(FSVLAFTER), !is.na(SITE.), 
         !is.na(Age), !is.na(FEMID.),
         Age >= 2)  # Adults only

# Check sample sizes
table(growth_data$Age, growth_data$SITE.)

# Separate by site
growth_data_OR <- growth_data %>% filter(SITE. == "OR")
growth_data_CP <- growth_data %>% filter(SITE. == "CP")
```

## Summary statistics
```{r}
# Summary statistics for growth data
summary_stats <- tibble(
  Population = c("Overall", "Warm lowland (OR)", "Cool highland (CP)"),
  Observations = c(nrow(growth_data), nrow(growth_data_OR), nrow(growth_data_CP)),
  Individuals = c(length(unique(growth_data$FEMID.)),
                  length(unique(growth_data_OR$FEMID.)),
                  length(unique(growth_data_CP$FEMID.))),
  Age_min = c(min(growth_data$Age), min(growth_data_OR$Age), min(growth_data_CP$Age)),
  Age_max = c(max(growth_data$Age), max(growth_data_OR$Age), max(growth_data_CP$Age)),
  Mean_obs_per_ind = c(
    round(mean(table(growth_data$FEMID.)), 1),
    round(mean(table(growth_data_OR$FEMID.)), 1),
    round(mean(table(growth_data_CP$FEMID.)), 1)),
  Obs_range_per_ind = c(
    paste0(min(table(growth_data$FEMID.)), "-", max(table(growth_data$FEMID.))),
    paste0(min(table(growth_data_OR$FEMID.)), "-", max(table(growth_data_OR$FEMID.))),
    paste0(min(table(growth_data_CP$FEMID.)), "-", max(table(growth_data_CP$FEMID.)))))

kable(summary_stats, digits = 1, 
      caption = "Summary statistics for growth data")

```

## Fit mixed effects von Bertalanffy models (with random effect on Linf)
```{r}
# ORFORD
fit_vb_OR <- nlme(
  FSVLAFTER ~ Linf * (1 - exp(-K * (Age - t0))),
  data = growth_data_OR,
  fixed = Linf + K + t0 ~ 1,
  random = Linf ~ 1 | FEMID.,
  start = c(Linf = 74, K = 0.044, t0 = 0),
  control = nlmeControl(maxIter = 200, msMaxIter = 200))

summary(fit_vb_OR)
intervals(fit_vb_OR)  # 95% CIs for fixed effects

# CENTRAL PLATEAU
fit_vb_CP <- nlme(
  FSVLAFTER ~ Linf * (1 - exp(-K * (Age - t0))),
  data = growth_data_CP,
  fixed = Linf + K + t0 ~ 1,
  random = Linf ~ 1 | FEMID.,
  start = c(Linf = 89, K = 0.044, t0 = 0),
  control = nlmeControl(maxIter = 200, msMaxIter = 200))

summary(fit_vb_CP)
intervals(fit_vb_CP)
```

## Model diagnostics
```{r}
# Function to check model diagnostics
check_nlme_diagnostics <- function(model, title) {
  
  par(mfrow = c(2, 2))
  
  # 1. Residuals vs Fitted
  plot(fitted(model), residuals(model, type = "pearson"),
       xlab = "Fitted values", ylab = "Pearson residuals",
       main = paste(title, "- Residuals vs Fitted"))
  abline(h = 0, col = "red", lty = 2)
  lines(lowess(fitted(model), residuals(model, type = "pearson")), 
        col = "blue", lwd = 2)
  
  # 2. QQ plot
  qqnorm(residuals(model, type = "pearson"), 
         main = paste(title, "- Normal Q-Q"))
  qqline(residuals(model, type = "pearson"), col = "red")
  
  # 3. Scale-Location
  plot(fitted(model), sqrt(abs(residuals(model, type = "pearson"))),
       xlab = "Fitted values", ylab = "√|Pearson residuals|",
       main = paste(title, "- Scale-Location"))
  lines(lowess(fitted(model), sqrt(abs(residuals(model, type = "pearson")))), 
        col = "blue", lwd = 2)
  
  # 4. Histogram of residuals
  hist(residuals(model, type = "pearson"), breaks = 20,
       main = paste(title, "- Residual Distribution"),
       xlab = "Pearson residuals")
  
  par(mfrow = c(1, 1))
  
  # Shapiro test
  cat("\n", title, "- Shapiro-Wilk test for normality:\n")
  print(shapiro.test(residuals(model, type = "pearson")))
}

# Check diagnostics for both models
check_nlme_diagnostics(fit_vb_OR, "OR")
check_nlme_diagnostics(fit_vb_CP, "CP")

# Random effects diagnostics
par(mfrow = c(1, 2))
qqnorm(ranef(fit_vb_OR)[,1], main = "OR - Random Effects Q-Q")
qqline(ranef(fit_vb_OR)[,1], col = "red")
qqnorm(ranef(fit_vb_CP)[,1], main = "CP - Random Effects Q-Q")
qqline(ranef(fit_vb_CP)[,1], col = "red")
par(mfrow = c(1, 1))

# good residual diagnostics
```

## Fit logistic and Gompertz models and compare with VB
```{r}
# Logistic OR
fit_log_OR <- nlme(
  FSVLAFTER ~ Linf / (1 + exp(-K * (Age - t0))),
  data = growth_data_OR,
  fixed = Linf + K + t0 ~ 1,
  random = Linf ~ 1 | FEMID.,
  start = c(Linf = 74, K = 0.5, t0 = 1),
  control = nlmeControl(maxIter = 200, msMaxIter = 200))

# Logistic CP
fit_log_CP <- nlme(
  FSVLAFTER ~ Linf / (1 + exp(-K * (Age - t0))),
  data = growth_data_CP,
  fixed = Linf + K + t0 ~ 1,
  random = Linf ~ 1 | FEMID.,
  start = c(Linf = 89, K = 0.5, t0 = 1),
  control = nlmeControl(maxIter = 200, msMaxIter = 200))

# Gompertz OR
fit_gomp_OR <- nlme(
  FSVLAFTER ~ Linf * exp(-exp(-K * (Age - t0))),
  data = growth_data_OR,
  fixed = Linf + K + t0 ~ 1,
  random = Linf ~ 1 | FEMID.,
  start = c(Linf = 74, K = 0.5, t0 = 0),
  control = nlmeControl(maxIter = 200, msMaxIter = 200))

# Gompertz CP
fit_gomp_CP <- nlme(
  FSVLAFTER ~ Linf * exp(-exp(-K * (Age - t0))),
  data = growth_data_CP,
  fixed = Linf + K + t0 ~ 1,
  random = Linf ~ 1 | FEMID.,
  start = c(Linf = 89, K = 0.5, t0 = 0),
  control = nlmeControl(maxIter = 200, msMaxIter = 200))

# Compare AICs
AIC(fit_vb_OR, fit_log_OR, fit_gomp_OR)
AIC(fit_vb_CP, fit_log_CP, fit_gomp_CP)

# VB fits both sites best (not much difference to others)
```

## Site comparison (combined model)
```{r}
# Combine datasets
growth_data_both <- growth_data %>%
  mutate(Site = factor(SITE.))

# Fit model with site-specific parameters
fit_combined <- nlme(
  FSVLAFTER ~ Linf * (1 - exp(-K * (Age - t0))),
  data = growth_data_both,
  fixed = list(Linf ~ Site, K ~ Site, t0 ~ Site),
  random = Linf ~ 1 | FEMID.,
  start = c(74, 15, 0.044, 0, 0, 0),  # Linf_OR, diff_CP, K_OR, diff_CP, t0_OR, diff_CP
  control = nlmeControl(maxIter = 200, msMaxIter = 200))

summary(fit_combined)

# Test if site effects are significant
anova(fit_combined)

## CP does have higher growth rate (K) 
## plus larger asymptote
# sligth difference between t0 

# so CP has faster approach to asymptote than OR

```

## Visualise
```{r}

# Generate prediction data
newdata_OR <- data.frame(Age = seq(2, 14, length.out = 100))
newdata_CP <- data.frame(Age = seq(2, 14, length.out = 100))

# Predict from models (population-level, fixed effects only)
pred_OR <- newdata_OR %>%
  mutate(
    FSVLAFTER = predict(fit_vb_OR, newdata = newdata_OR, level = 0),
    Site = "OR")

pred_CP <- newdata_CP %>%
  mutate(
    FSVLAFTER = predict(fit_vb_CP, newdata = newdata_CP, level = 0),
    Site = "CP")

# For confidence intervals, use parametric bootstrap
set.seed(123)

# Bootstrap OR
boot_predict_OR <- function() {
  # Simulate new parameters from model
  beta <- fixef(fit_vb_OR)
  vcov_beta <- vcov(fit_vb_OR)
  sim_beta <- MASS::mvrnorm(1, beta, vcov_beta)
  
  # Predict with simulated parameters
  Linf_sim <- sim_beta[1]
  K_sim <- sim_beta[2]
  t0_sim <- sim_beta[3]
  
  Linf_sim * (1 - exp(-K_sim * (newdata_OR$Age - t0_sim)))
}

boot_OR <- replicate(1000, boot_predict_OR())
pred_OR$lower <- apply(boot_OR, 1, quantile, 0.025)
pred_OR$upper <- apply(boot_OR, 1, quantile, 0.975)

# Bootstrap CP
boot_predict_CP <- function() {
  beta <- fixef(fit_vb_CP)
  vcov_beta <- vcov(fit_vb_CP)
  sim_beta <- MASS::mvrnorm(1, beta, vcov_beta)
  
  Linf_sim <- sim_beta[1]
  K_sim <- sim_beta[2]
  t0_sim <- sim_beta[3]
  
  Linf_sim * (1 - exp(-K_sim * (newdata_CP$Age - t0_sim)))
}

boot_CP <- replicate(1000, boot_predict_CP())
pred_CP$lower <- apply(boot_CP, 1, quantile, 0.025)
pred_CP$upper <- apply(boot_CP, 1, quantile, 0.975)

# Combine predictions
pred_all <- bind_rows(pred_OR, pred_CP)


#FINAL PUBLICATION-QUALITY PLOT

# Prepare data for plotting
growth_data_plot <- growth_data %>%
  mutate(Site = factor(SITE., levels = c("OR", "CP")))

# Extract Linf values for horizontal lines
Linf_OR <- fixef(fit_vb_OR)["Linf"]
Linf_CP <- fixef(fit_vb_CP)["Linf"]

# Create plot
p_growth <- ggplot() +
    geom_jitter(data = growth_data_plot, 
              aes(x = Age, y = FSVLAFTER, colour = Site),
              width = 0.15, height = 0, alpha = 0.3, size = 1.5) +
    geom_line(data = pred_all, 
            aes(x = Age, y = FSVLAFTER, colour = Site), 
            linewidth = 1.2) +
    geom_ribbon(data = pred_all, 
              aes(x = Age, ymin = lower, ymax = upper, fill = Site), 
              alpha = 0.2) +
    geom_hline(yintercept = Linf_OR, 
             colour = "coral1", linetype = "dashed", linewidth = 0.6) +
  geom_hline(yintercept = Linf_CP, 
             colour = "cornflowerblue", linetype = "dashed", linewidth = 0.6) +
    scale_x_continuous(breaks = seq(2, 14, by = 2), limits = c(2, 14)) +
  scale_colour_manual(values = c("OR" = "coral1", "CP" = "cornflowerblue")) +
  scale_fill_manual(values = c("OR" = "coral1", "CP" = "cornflowerblue")) +
    labs(x = "Age (years)", 
       y = "Snout-vent length (mm)") +
  theme_classic(base_size = 12) +
  theme(legend.position = "none")

print(p_growth)


ggsave("GrowthCurve.pdf", p_growth, 
       width = 18, height = 13, units = "cm", dpi = 600)

ggsave("GrowthCurve.png", p_growth, 
       width = 18, height = 13, units = "cm", dpi = 600)

```


## Results table (supplementary material)
```{r}
results_table <- tibble(
  Population = rep(c("Warm lowland (OR)", "Cool highland (CP)"), each = 3),
  Parameter = rep(c("L∞ (mm)", "K (year⁻¹)", "t₀ (year)"), 2),
  Estimate = c(fixef(fit_vb_OR), fixef(fit_vb_CP)),
  Lower_95CI = c(intervals(fit_vb_OR)$fixed[,"lower"], intervals(fit_vb_CP)$fixed[,"lower"]),
  Upper_95CI = c(intervals(fit_vb_OR)$fixed[,"upper"], intervals(fit_vb_CP)$fixed[,"upper"])) %>%
  mutate(across(c(Estimate, Lower_95CI, Upper_95CI), ~round(., 3)))

kable(results_table, caption = "Growth parameters and 95% CIs for each population")
```



## Extract key statistics for manuscript
```{r}
manuscript_stats <- tibble(
  Population = c("Warm lowland (OR)", "Cool highland (CP)"),
  Observations = c(nrow(growth_data_OR), nrow(growth_data_CP)),
  Individuals = c(length(unique(growth_data_OR$FEMID.)), length(unique(growth_data_CP$FEMID.))),
  SD_Random_Linf = c(as.numeric(VarCorr(fit_vb_OR)[1,2]), as.numeric(VarCorr(fit_vb_CP)[1,2])),
  Residual_SD = c(fit_vb_OR$sigma, fit_vb_CP$sigma))

kable(manuscript_stats, digits = 2, caption = "Sample sizes and model statistics for manuscript")

```


## Early growth comparison
```{r}
## Early growth comparison (ages 0-2 years) - pre-reproductive phase

growth_data_all <- read.csv("Growthcurvedata.csv", 
                        na.strings = c("", ",", "NA", "."), 
                        header = TRUE)

# Subset to early ages
early_growth <- growth_data_all %>%
  filter(!is.na(FSVLAFTER), !is.na(SITE.), 
         !is.na(Age), 
         Age <= 2) %>%
  mutate(Site = factor(SITE., levels = c("OR", "CP")))

# Check sample sizes
table(early_growth$Age, early_growth$Site)

# Linear model with interaction to test if growth rates differ
early_model <- lm(FSVLAFTER ~ Age * Site, data = early_growth)
summary(early_model)

# Extract coefficients
coef_summary <- summary(early_model)$coefficients
cat("\nInterpretation:\n")
cat("OR growth rate (mm/year):", round(coef_summary["Age", "Estimate"], 2), "\n")
cat("CP growth rate (mm/year):", round(coef_summary["Age", "Estimate"] + coef_summary["Age:SiteCP", "Estimate"], 2), "\n")
cat("Difference:", round(coef_summary["Age:SiteCP", "Estimate"], 2), "\n")
cat("P-value for difference:", round(coef_summary["Age:SiteCP", "Pr(>|t|)"], 4), "\n")

# Visualize
ggplot(early_growth, aes(x = Age, y = FSVLAFTER, color = Site)) +
  geom_jitter(alpha = 0.3, width = 0.02) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.5) +
  scale_color_manual(values = c("OR" = "coral1", "CP" = "cornflowerblue")) +
  labs(x = "Age (years)", y = "SVL (mm)", 
       title = "Early growth rates (ages 0-2)") +
  theme_classic(base_size = 12)

# early growth rates also differ

# check quadratic
# Compare linear vs polynomial
early_model_quad <- lm(FSVLAFTER ~ poly(Age, 2) * Site, data = early_growth)
AIC(early_model, early_model_quad)

# Check the quadratic model
summary(early_model_quad)

# Visualize to see the curvature
ggplot(early_growth, aes(x = Age, y = FSVLAFTER, color = Site)) +
  geom_jitter(alpha = 0.3, width = 0.02) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), 
              se = TRUE, linewidth = 1.5) +
  scale_color_manual(values = c("OR" = "coral1", "CP" = "cornflowerblue")) +
  labs(x = "Age (years)", y = "SVL (mm)", 
       title = "Early growth: quadratic model (ages 0-2)") +
  theme_classic(base_size = 12)

summary(early_model_quad)

# Check specific interactions
anova(early_model_quad)


# Linear interaction (poly(Age, 2)1:SiteCP): p < 2e-16 - CP's overall growth trajectory differs
# Quadratic interaction (poly(Age, 2)2:SiteCP): p < 2e-16 - CP's curvature pattern differs
# R² improved from 0.92 (linear) to 0.95 (quadratic)
# CP grow faster AND they have a different growth acceleration pattern during ages 0-2.

```


## Visualise all ages for publication
### von Bertalanffy growth curves fitted to post-reproductive individuals (age ≥2 years). Raw data shown for all ages including juveniles.
```{r}
## Publication figure with all ages shown

# Generate predictions from nlme models (ages 2+)
newdata_OR <- data.frame(Age = seq(2, 14, length.out = 100))
newdata_CP <- data.frame(Age = seq(2, 14, length.out = 100))

# Predict from models (population-level, fixed effects only)
pred_OR <- newdata_OR %>%
  mutate(
    FSVLAFTER = predict(fit_vb_OR, newdata = newdata_OR, level = 0),
    Site = "OR")

pred_CP <- newdata_CP %>%
  mutate(
    FSVLAFTER = predict(fit_vb_CP, newdata = newdata_CP, level = 0),
    Site = "CP")

# Bootstrap for confidence intervals
set.seed(123)

boot_predict_OR <- function() {
  beta <- fixef(fit_vb_OR)
  vcov_beta <- vcov(fit_vb_OR)
  sim_beta <- MASS::mvrnorm(1, beta, vcov_beta)
  
  Linf_sim <- sim_beta[1]
  K_sim <- sim_beta[2]
  t0_sim <- sim_beta[3]
  
  Linf_sim * (1 - exp(-K_sim * (newdata_OR$Age - t0_sim)))
}

boot_OR <- replicate(1000, boot_predict_OR())
pred_OR$lower <- apply(boot_OR, 1, quantile, 0.025)
pred_OR$upper <- apply(boot_OR, 1, quantile, 0.975)

boot_predict_CP <- function() {
  beta <- fixef(fit_vb_CP)
  vcov_beta <- vcov(fit_vb_CP)
  sim_beta <- MASS::mvrnorm(1, beta, vcov_beta)
  
  Linf_sim <- sim_beta[1]
  K_sim <- sim_beta[2]
  t0_sim <- sim_beta[3]
  
  Linf_sim * (1 - exp(-K_sim * (newdata_CP$Age - t0_sim)))
}

boot_CP <- replicate(1000, boot_predict_CP())
pred_CP$lower <- apply(boot_CP, 1, quantile, 0.025)
pred_CP$upper <- apply(boot_CP, 1, quantile, 0.975)

# Combine predictions
pred_all <- bind_rows(pred_OR, pred_CP)

# Extract Linf values for horizontal lines
Linf_OR <- fixef(fit_vb_OR)["Linf"]
Linf_CP <- fixef(fit_vb_CP)["Linf"]

# Prepare ALL growth data for plotting (ages 0-14)
growth_data_plot <- growth_data_all %>%
  mutate(Site = factor(SITE., levels = c("OR", "CP")))

# Create plot
p_growth <- ggplot() +
  # Raw data points - ALL AGES
  geom_jitter(data = growth_data_plot, 
              aes(x = Age, y = FSVLAFTER, colour = Site),
              width = 0.15, height = 0, alpha = 0.3, size = 1.5) +
  # Fitted curves - AGES 2+ ONLY
  geom_line(data = pred_all, 
            aes(x = Age, y = FSVLAFTER, colour = Site), 
            linewidth = 1.2) +
  geom_ribbon(data = pred_all, 
              aes(x = Age, ymin = lower, ymax = upper, fill = Site), 
              alpha = 0.2) +
  # Asymptote lines
  geom_hline(yintercept = Linf_OR, 
             colour = "coral1", linetype = "dashed", linewidth = 0.6) +
  geom_hline(yintercept = Linf_CP, 
             colour = "cornflowerblue", linetype = "dashed", linewidth = 0.6) +
  scale_x_continuous(breaks = seq(0, 14, by = 2), limits = c(0, 14)) +
  scale_colour_manual(values = c("OR" = "coral1", "CP" = "cornflowerblue")) +
  scale_fill_manual(values = c("OR" = "coral1", "CP" = "cornflowerblue")) +
  labs(x = "Age (years)", 
       y = "Snout-vent length (mm)") +
  theme_classic(base_size = 12) +
  theme(legend.position = "none")

print(p_growth)


## Add quadratic predictions for ages 0-2 to the growth curve plot

# Generate predictions from quadratic model for early ages
early_pred <- expand.grid(
  Age = seq(0, 2, length.out = 100),
  Site = factor(c("OR", "CP"), levels = c("OR", "CP"))
)

# Predict with confidence intervals
early_pred <- early_pred %>%
  mutate(
    FSVLAFTER = predict(early_model_quad, newdata = early_pred),
    # Get prediction intervals
    pred_interval = predict(early_model, newdata = early_pred, 
                           interval = "confidence", level = 0.95)
  )

# Extract confidence intervals properly
pred_with_ci <- predict(early_model_quad, newdata = early_pred, 
                       interval = "confidence", level = 0.95)
early_pred$lower <- pred_with_ci[, "lwr"]
early_pred$upper <- pred_with_ci[, "upr"]
early_pred$pred_interval <- NULL

# add to plot
p_growth <- ggplot() +
  # Raw data - ALL AGES
  geom_jitter(data = growth_data_plot, 
              aes(x = Age, y = FSVLAFTER, colour = Site),
              width = 0.15, height = 0, alpha = 0.3, size = 1.5) +
  # LINEAR predictions for ages 0-2
  geom_line(data = early_pred, 
            aes(x = Age, y = FSVLAFTER, colour = Site), 
            linewidth = 1.2) +
  geom_ribbon(data = early_pred, 
              aes(x = Age, ymin = lower, ymax = upper, fill = Site), 
              alpha = 0.2) +
  # VON BERTALANFFY predictions for ages 2+
  geom_line(data = pred_all, 
            aes(x = Age, y = FSVLAFTER, colour = Site), 
            linewidth = 1.2) +
  geom_ribbon(data = pred_all, 
              aes(x = Age, ymin = lower, ymax = upper, fill = Site), 
              alpha = 0.2) +
  # Asymptote lines
  geom_hline(yintercept = Linf_OR, 
             colour = "coral1", linetype = "dashed", linewidth = 0.6) +
  geom_hline(yintercept = Linf_CP, 
             colour = "cornflowerblue", linetype = "dashed", linewidth = 0.6) +
  scale_x_continuous(breaks = seq(0, 14, by = 2), limits = c(0, 14)) +
  scale_colour_manual(values = c("OR" = "coral1", "CP" = "cornflowerblue")) +
  scale_fill_manual(values = c("OR" = "coral1", "CP" = "cornflowerblue")) +
  labs(x = "Age (years)", y = "Snout-vent length (mm)") +
  theme_classic(base_size = 12) +
  theme(legend.position = "none")

p_growth

ggsave("GrowthCurve_final.pdf", p_growth, 
       width = 10, height = 10, units = "cm", dpi = 600)
ggsave("GrowthCurve_final.png", p_growth, 
       width = 10, height = 10, units = "cm", dpi = 600)



```


