---
title: "Growth Curve Analysis 2025"
author: "Luisa Fitzpatrick"
date: "r Sys.Date()"
output:
  word_document: default
  html_document: default
---

```{r load-packages, message = FALSE, warning = FALSE}
library(nlme)
library(ggplot2)
library(knitr)
library(dplyr)
# Note: MASS is used via MASS::mvrnorm() in the bootstrap functions
```


## Results table (supplementary material)
```{r}
results_table <- tibble(
  Population = rep(c("Warm lowland (OR)", "Cool highland (CP)"), each = 3),
  Parameter = rep(c("L∞ (mm)", "K (year⁻¹)", "t₀ (year)"), 2),
  Estimate = c(fixef(fit_vb_OR), fixef(fit_vb_CP)),
  Lower_95CI = c(intervals(fit_vb_OR)$fixed[,"lower"], intervals(fit_vb_CP)$fixed[,"lower"]),
  Upper_95CI = c(intervals(fit_vb_OR)$fixed[,"upper"], intervals(fit_vb_CP)$fixed[,"upper"])) %>%
  mutate(across(c(Estimate, Lower_95CI, Upper_95CI), ~round(., 3)))

kable(results_table, caption = "Growth parameters and 95% CIs for each population")
```



## Extract key statistics for manuscript
```{r}
manuscript_stats <- tibble(
  Population = c("Warm lowland (OR)", "Cool highland (CP)"),
  Observations = c(nrow(growth_data_OR), nrow(growth_data_CP)),
  Individuals = c(length(unique(growth_data_OR$FEMID.)), length(unique(growth_data_CP$FEMID.))),
  SD_Random_Linf = c(as.numeric(VarCorr(fit_vb_OR)[1,2]), as.numeric(VarCorr(fit_vb_CP)[1,2])),
  Residual_SD = c(fit_vb_OR$sigma, fit_vb_CP$sigma))

kable(manuscript_stats, digits = 2, caption = "Sample sizes and model statistics for manuscript")

```


## Including all female offspring and juvenile samples to test early growth differences
#### Data Preparation

```{r}
# Read in the data
growth_data_all <- read.csv("Growthcurvedata.csv", na.strings = c("", ",", "NA", "."), header = TRUE)

# Separate by site
growth_data_all_OR <- growth_data_all %>% filter(SITE. == "OR")
growth_data_all_CP <- growth_data_all %>% filter(SITE. == "CP")

```

### 1. Compare size at birth between sites (cross-sectional)
```{r}
birth_data <- growth_data_all %>% filter(Age == 0.01)
t.test(FSVLAFTER ~ SITE., data = birth_data)

# t test resulted in significant difference between size at birth 
# OR mean 29.63, CP mean 30.20
# t = 19.463, df = 3621.2, p-value < 2.2e-16
```

### 2. Compare juvenile sizes between sites (cross-sectional)
```{r}
#Compare birth sizes between sites (cross-sectional)
juv_data <- growth_data_all %>% filter(Age == 0.75)
t.test(FSVLAFTER ~ SITE., data = juv_data)

# t test resulted in significant difference between size after first winter 
# OR mean 37.26, CP mean 38.27
# t = 2.2975, df = 179.92, p-value = 0.02274
```

### Visualise Raw Data (with birth size and juvenile size)
```{r}
ggplot(growth_data_all, aes(x = Age, y = FSVLAFTER, colour = SITE.)) +
  geom_jitter(alpha = 0.3, width = 0.2, height = 0) +
  scale_colour_manual(values = c("OR" = "coral1", "CP" = "cornflowerblue")) +
  theme_classic() +
  labs(title = "Raw SVL vs Age by Population")
```

## Summary statistics
```{r}
# Grouped age summary
age_class_summary <- growth_data_all %>%
  mutate(Age_class = case_when(
    Age == 0.01 ~ "Offspring (0.01)",
    Age == 0.75 ~ "Age 0.75",
    Age >= 1 & Age < 2 ~ "Age 1",
    Age >= 2 & Age <= 5 ~ "Age 2-5",
    Age > 5 ~ "Age 6+")) %>%
  group_by(SITE., Age_class) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = SITE., values_from = n, values_fill = 0) %>%
  mutate(Total = CP + OR)

kable(age_class_summary, 
      caption = "Sample sizes by age class and population")

```

## Fit von Bertalanffy models 
```{r}

# ORFORD
fit_vb_OR <- nls(
  FSVLAFTER ~ Linf * (1 - exp(-K * (Age - t0))),
  data = growth_data_all_OR,
  start = c(Linf = 74, K = 0.044, t0 = 0))

summary(fit_vb_OR)
confint(fit_vb_OR)  # 95% CIs for parameters

# CENTRAL PLATEAU
fit_vb_CP <- nls(
  FSVLAFTER ~ Linf * (1 - exp(-K * (Age - t0))),
  data = growth_data_all_CP,
  start = c(Linf = 89, K = 0.044, t0 = 0))

summary(fit_vb_CP)
confint(fit_vb_CP)
```

## CHeck VB model diagnostics
```{r}
check_nls_diagnostics <- function(model, title) {
  
  par(mfrow = c(2, 2))
  
  # 1. Residuals vs Fitted
  plot(fitted(model), residuals(model),
       xlab = "Fitted values", ylab = "Residuals",
       main = paste(title, "- Residuals vs Fitted"))
  abline(h = 0, col = "red", lty = 2)
  lines(lowess(fitted(model), residuals(model)), 
        col = "blue", lwd = 2)
  
  # 2. QQ plot
  qqnorm(residuals(model), 
         main = paste(title, "- Normal Q-Q"))
  qqline(residuals(model), col = "red")
  
  # 3. Scale-Location
  plot(fitted(model), sqrt(abs(residuals(model))),
       xlab = "Fitted values", ylab = "√|Residuals|",
       main = paste(title, "- Scale-Location"))
  lines(lowess(fitted(model), sqrt(abs(residuals(model)))), 
        col = "blue", lwd = 2)
  
  # 4. Histogram of residuals
  hist(residuals(model), breaks = 20,
       main = paste(title, "- Residual Distribution"),
       xlab = "Residuals")
  
  par(mfrow = c(1, 1))
  
  # Shapiro test
  cat("\n", title, "- Shapiro-Wilk test for normality:\n")
  print(shapiro.test(residuals(model)))
}

# Check diagnostics for both models
check_nls_diagnostics(fit_vb_OR, "OR")
check_nls_diagnostics(fit_vb_CP, "CP")

# Try weighted regressions, with variance proportional to size 

## Fit von Bertalanffy models with weights to address heteroscedasticity

# ORFORD - weighted
fit_vb_OR <- nls(
  FSVLAFTER ~ Linf * (1 - exp(-K * (Age - t0))),
  data = growth_data_all_OR,
  start = c(Linf = 74, K = 0.044, t0 = 0),
  weights = 1/FSVLAFTER)

summary(fit_vb_OR)
confint(fit_vb_OR)

# CENTRAL PLATEAU - weighted
fit_vb_CP <- nls(
  FSVLAFTER ~ Linf * (1 - exp(-K * (Age - t0))),
  data = growth_data_all_CP,
  start = c(Linf = 89, K = 0.044, t0 = 0),
  weights = 1/FSVLAFTER)

summary(fit_vb_CP)
confint(fit_vb_CP)

# Check diagnostics
check_nls_diagnostics(fit_vb_OR, "OR - Weighted")
check_nls_diagnostics(fit_vb_CP, "CP - Weighted")

# no imporvement

```

## Fit logistic and Gompertz models and compare with VB
```{r}
# Logistic OR
fit_log_OR <- nls(
  FSVLAFTER ~ Linf / (1 + exp(-K * (Age - t0))),
  data = growth_data_all_OR,
  start = c(Linf = 74, K = 0.5, t0 = 1))

# Logistic CP
fit_log_CP <- nls(
  FSVLAFTER ~ Linf / (1 + exp(-K * (Age - t0))),
  data = growth_data_all_CP,
  start = c(Linf = 89, K = 0.5, t0 = 1))

# Gompertz OR
fit_gomp_OR <- nls(
  FSVLAFTER ~ Linf * exp(-exp(-K * (Age - t0))),
  data = growth_data_all_OR,
  start = c(Linf = 74, K = 0.5, t0 = 0))

# Gompertz CP
fit_gomp_CP <- nls(
  FSVLAFTER ~ Linf * exp(-exp(-K * (Age - t0))),
  data = growth_data_all_CP,
  start = c(Linf = 89, K = 0.5, t0 = 0))

# Compare AICs
cat("\nOR Models:\n")
AIC(fit_vb_OR, fit_log_OR, fit_gomp_OR)

cat("\nCP Models:\n")
AIC(fit_vb_CP, fit_log_CP, fit_gomp_CP)

# VB is best for OR
# Logistic is slightly better for CP, but is this just driven by early steep rise and not accounting for late plateau

## Compare VB vs Logistic fits for CP visually

# Generate predictions across full age range
age_seq <- seq(0, 14, length.out = 200)

pred_vb_CP <- data.frame(
  Age = age_seq,
  SVL = predict(fit_vb_CP, newdata = data.frame(Age = age_seq)),
  Model = "von Bertalanffy"
)

pred_log_CP <- data.frame(
  Age = age_seq,
  SVL = predict(fit_log_CP, newdata = data.frame(Age = age_seq)),
  Model = "Logistic"
)

# Plot both
ggplot() +
  geom_jitter(data = growth_data_all_CP, 
              aes(x = Age, y = FSVLAFTER), 
              alpha = 0.3, width = 0.15, color = "cornflowerblue") +
  geom_line(data = bind_rows(pred_vb_CP, pred_log_CP),
            aes(x = Age, y = SVL, color = Model, linetype = Model),
            linewidth = 1.2) +
  scale_color_manual(values = c("von Bertalanffy" = "blue", "Logistic" = "red")) +
  labs(title = "CP: von Bertalanffy vs Logistic",
       x = "Age (years)", y = "SVL (mm)") +
  theme_classic()

# Compare Linf estimates
cat("VB Linf:", coef(fit_vb_CP)["Linf"], "\n")
cat("Logistic Linf:", coef(fit_log_CP)["Linf"], "\n")

#logistic is underestimating the ultimate asymptotic size because it's forced to plateau earlier, but it better fits the early steep growth pattern. 

```

## Site comparison (combined model)
```{r}
# Prepare data
growth_data_both <- bind_rows(
  growth_data_all_OR %>% mutate(Site = "OR"),
  growth_data_all_CP %>% mutate(Site = "CP")
) %>%
  mutate(Site = factor(Site, levels = c("OR", "CP")))

# Fit model with site-specific parameters using indexing
fit_combined <- nls(
  FSVLAFTER ~ Linf[Site] * (1 - exp(-K[Site] * (Age - t0[Site]))),
  data = growth_data_both,
  start = list(Linf = c(74, 89), 
               K = c(0.044, 0.044), 
               t0 = c(0, 0)))

summary(fit_combined)
confint(fit_combined)



# Test if K differs between sites
fit_common_K <- nls(
  FSVLAFTER ~ Linf[Site] * (1 - exp(-K * (Age - t0[Site]))),
  data = growth_data_both,
  start = list(Linf = c(74, 89), K = 0.044, t0 = c(0, 0)))

# Compare models
anova(fit_common_K, fit_combined)


# no difference in growth rate between CP and OR, CP just keep growing for longer
# however, i think there is a difference in growth rate during the first two years

## Compare predicted early growth using fitted VB parameters

# Extract parameters from combined model
params <- coef(fit_combined)

# Function to predict size at age from VB equation
predict_size <- function(age, Linf, K, t0) {
  Linf * (1 - exp(-K * (age - t0)))
}

# Calculate predicted sizes at ages 0 and 3 for each site
OR_size_0 <- predict_size(0, params["Linf1"], params["K1"], params["t01"])
OR_size_3 <- predict_size(3, params["Linf1"], params["K1"], params["t01"])
OR_growth_0to3 <- OR_size_3 - OR_size_0

CP_size_0 <- predict_size(0, params["Linf2"], params["K2"], params["t02"])
CP_size_3 <- predict_size(3, params["Linf2"], params["K2"], params["t02"])
CP_growth_0to3 <- CP_size_3 - CP_size_0

# Compare
cat("Predicted growth from age 0 to 3:\n")
cat("OR:", round(OR_growth_0to3, 2), "mm\n")
cat("CP:", round(CP_growth_0to3, 2), "mm\n")
cat("Difference:", round(CP_growth_0to3 - OR_growth_0to3, 2), "mm\n")

## Compare predicted sizes across early ages
ages <- c(0, 0.01, 0.75, 1, 2, 3)

growth_trajectory <- data.frame(
  Age = rep(ages, 2),
  Site = rep(c("OR", "CP"), each = length(ages))
) %>%
  mutate(
    Predicted_size = ifelse(Site == "OR",
                           predict_size(Age, params["Linf1"], params["K1"], params["t01"]),
                           predict_size(Age, params["Linf2"], params["K2"], params["t02"]))
  )

# View the trajectory
growth_trajectory %>%
  pivot_wider(names_from = Site, values_from = Predicted_size) %>%
  mutate(Difference = CP - OR)

# Visualize
ggplot(growth_trajectory, aes(x = Age, y = Predicted_size, color = Site)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  labs(y = "Predicted SVL (mm)", x = "Age (years)",
       title = "Predicted early growth trajectories")

```


## Visualise
```{r}
# Generate prediction data - now including age 0 to show early growth
newdata <- data.frame(
  Age = seq(0, 14, length.out = 200),
  Site = rep(c("OR", "CP"), each = 200)
) %>%
  mutate(Site = factor(Site, levels = c("OR", "CP")))

# Predict from combined model
pred_all <- newdata %>%
  mutate(FSVLAFTER = predict(fit_combined, newdata = .))

# For confidence intervals, use parametric bootstrap
set.seed(123)

# Bootstrap predictions
boot_predict <- function() {
  # Simulate new parameters from model
  beta <- coef(fit_combined)
  vcov_beta <- vcov(fit_combined)
  sim_beta <- MASS::mvrnorm(1, beta, vcov_beta)
  
  # Predict with simulated parameters for each site
  predictions <- numeric(nrow(newdata))
  
  for(i in 1:nrow(newdata)) {
    site_idx <- ifelse(newdata$Site[i] == "OR", 1, 2)
    Linf_sim <- sim_beta[site_idx]
    K_sim <- sim_beta[site_idx + 2]
    t0_sim <- sim_beta[site_idx + 4]
    
    predictions[i] <- Linf_sim * (1 - exp(-K_sim * (newdata$Age[i] - t0_sim)))
  }
  
  return(predictions)
}

boot_results <- replicate(1000, boot_predict())
pred_all$lower <- apply(boot_results, 1, quantile, 0.025)
pred_all$upper <- apply(boot_results, 1, quantile, 0.975)

# Extract Linf values for horizontal lines
params <- coef(fit_combined)
Linf_OR <- params["Linf1"]
Linf_CP <- params["Linf2"]

# Prepare data for plotting
growth_data_plot <- growth_data_both %>%
  mutate(Site = factor(Site, levels = c("OR", "CP")))

# Create plot
p_growth <- ggplot() +
  geom_jitter(data = growth_data_plot, 
              aes(x = Age, y = FSVLAFTER, colour = Site),
              width = 0.15, height = 0, alpha = 0.3, size = 1.5) +
  geom_line(data = pred_all, 
            aes(x = Age, y = FSVLAFTER, colour = Site), 
            linewidth = 1.2) +
  geom_ribbon(data = pred_all, 
              aes(x = Age, ymin = lower, ymax = upper, fill = Site), 
              alpha = 0.2) +
  geom_hline(yintercept = Linf_OR, 
             colour = "coral1", linetype = "dashed", linewidth = 0.6) +
  geom_hline(yintercept = Linf_CP, 
             colour = "cornflowerblue", linetype = "dashed", linewidth = 0.6) +
  scale_x_continuous(breaks = seq(0, 14, by = 2), limits = c(0, 14)) +
  scale_colour_manual(values = c("OR" = "coral1", "CP" = "cornflowerblue")) +
  scale_fill_manual(values = c("OR" = "coral1", "CP" = "cornflowerblue")) +
  labs(x = "Age (years)", 
       y = "Snout-vent length (mm)") +
  theme_classic(base_size = 12) +
  theme(legend.position = "none")

print(p_growth)

ggsave("GrowthCurveall.pdf", p_growth, 
       width = 18, height = 13, units = "cm", dpi = 600)
ggsave("GrowthCurveall.png", p_growth, 
       width = 18, height = 13, units = "cm", dpi = 600)

```