---
title: "Manuscript Figures — Ocellatus Senescence"
author: "Luisa Fitzpatrick"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 6)
```

```{r libraries, message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(cowplot)
library(ggplot2)
library(nlme)
library(MASS)
library(BaSTA)
library(BaFTA)
library(rnaturalearth)
library(terra)
library(sf)
library(geodata)
library(magick)
```

---

# Figure 1 — Study System

**Required files:**

- `data/processed/BOM.csv`
- `images/ocellatus.jpg`

```{r figure1-data}

# ── Month label helper (single letters for x-axes) ──────────────────────────
month_letters <- c("J","F","M","A","M","J","J","A","S","O","N","D")

# ── Shared theme ─────────────────────────────────────────────────────────────
theme_fig1 <- theme_classic(base_size = 11) +
  theme(axis.text  = element_text(color = "black"),
        axis.line  = element_line(color = "black"),
        plot.margin = margin(12, 8, 5, 8))   # extra top margin for panel labels

site_colors <- c("OR" = "coral1", "CP" = "cornflowerblue")
site_labels <- c("OR" = "Lowland", "CP" = "Highland")
```

## Panel A — Species photo

```{r fig1-panelA}
photo_species <- magick::image_read(here("images", "ocellatus.jpg"))
p_species <- ggdraw() + draw_image(photo_species)
```

## Panel B — Tasmania elevation map

```{r fig1-panelB, message=FALSE}
# Download elevation (cached to tempdir after first run)
elev     <- elevation_30s(country = "AUS", path = tempdir())
tas_ext  <- ext(144.5, 148.5, -43.7, -40.0)
elev_tas <- crop(elev, tas_ext)

elev_df  <- as.data.frame(elev_tas, xy = TRUE)
colnames(elev_df) <- c("lon", "lat", "elevation")
elev_df  <- elev_df[!is.na(elev_df$elevation), ]

sites <- data.frame(
  name = c("Highland", "Lowland"),
  lon  = c(146.5722, 147.8489),
  lat  = c(-41.8447, -42.5594))

map1 <- ggplot() +
  geom_raster(data = elev_df, aes(x = lon, y = lat, fill = elevation)) +
  scale_fill_gradientn(
    colours = terrain.colors(10),
    name    = "Elevation (m)") +
  geom_point(data = sites, aes(x = lon, y = lat),
             shape = 21, fill = c("cornflowerblue", "coral1"),
             color = "black", size = 5, stroke = 1.2) +
  geom_text(data = sites, aes(x = lon, y = lat, label = name),
            hjust = 0.6, vjust = 2.4, size = 3.5, fontface = "bold") +
  coord_sf(xlim = c(144.5, 148.5), ylim = c(-43.7, -40.0),
           crs = 4326, expand = FALSE, datum = sf::st_crs(4326)) +
  theme_void(base_size = 11) +
  theme(
    # Extra margins so the elevation legend is never clipped
    plot.margin          = margin(12, 8, 15, 8),
    legend.position      = "bottom",
    legend.direction     = "horizontal",
    legend.background    = element_rect(fill = "white", color = NA),
    legend.title         = element_text(size = 10),
    legend.text          = element_text(size = 9),
    legend.key.width     = unit(1.0, "cm"),
    legend.key.height    = unit(0.3, "cm"),
    legend.margin        = margin(4, 0, 0, 0),
    legend.box.margin    = margin(0, 0, 0, 0),
    plot.clip            = "off")

map1
```

## Panels C & D — Climate

```{r fig1-panelCD}
bom <- read.csv(here("data", "processed", "BOM.csv"))

monthly_summary <- bom %>%
  group_by(Site, Month) %>%
  summarise(
    AvgMaxTemp = mean(Maxtemp,  na.rm = TRUE),
    SDMaxTemp  = sd(Maxtemp,   na.rm = TRUE),
    AvgMinTemp = mean(Mintemp,  na.rm = TRUE),
    SDMinTemp  = sd(Mintemp,   na.rm = TRUE),
    TotalRain  = sum(Rainfall,  na.rm = TRUE),
    SDRain     = sd(Rainfall,   na.rm = TRUE) * sqrt(n()),
    .groups    = "drop") %>%
  mutate(Month = factor(Month, levels = 1:12, ordered = TRUE))

# Panel C — Temperature
temp_plot <- ggplot(monthly_summary,
                    aes(x = as.numeric(Month), group = Site,
                        color = Site, fill = Site)) +
  geom_ribbon(aes(ymin = AvgMaxTemp - SDMaxTemp,
                  ymax = AvgMaxTemp + SDMaxTemp),
              alpha = 0.2, color = NA) +
  geom_line(aes(y = AvgMaxTemp), linewidth = 1.2, linetype = "solid") +
  geom_point(aes(y = AvgMaxTemp), size = 2.5) +
  geom_ribbon(aes(ymin = AvgMinTemp - SDMinTemp,
                  ymax = AvgMinTemp + SDMinTemp),
              alpha = 0.2, color = NA) +
  geom_line(aes(y = AvgMinTemp), linewidth = 1.2, linetype = "dashed") +
  geom_point(aes(y = AvgMinTemp), size = 2.5, shape = 1) +
  scale_color_manual(values = site_colors, labels = site_labels) +
  scale_fill_manual(values  = site_colors, labels = site_labels) +
  scale_y_continuous(labels = function(x) paste0(x, "    ")) +
  scale_x_continuous(breaks = 1:12, labels = month_letters,
                   limits = c(0.5, 12.5), expand = expansion(add = 0)) +
  labs(y = "Mean daily temperature (°C)", x = NULL,
       color = "Population", fill = "Population") +
  theme_fig1 +
  theme(legend.position = "none")
  
  temp_plot

# Panel D — Rainfall
rain_plot <- ggplot(monthly_summary,
                    aes(x = as.numeric(Month), y = TotalRain, fill = Site)) +
  geom_col(aes(x = as.numeric(Month) + ifelse(Site == "OR", -0.2, 0.2)),
           position = "identity", width = 0.35) +
  geom_errorbar(aes(x   = as.numeric(Month) + ifelse(Site == "OR", -0.2, 0.2),
                    ymin = TotalRain,
                    ymax = TotalRain + SDRain),
                width = 0.15, linewidth = 0.5) +
  scale_fill_manual(values = site_colors, labels = site_labels) +
  scale_x_continuous(breaks = 1:12, labels = month_letters,
                   limits = c(0.5, 12.5), expand = expansion(add = 0)) +
  scale_y_continuous(
    expand = c(0, 0),
    limits = c(0, max(monthly_summary$TotalRain + monthly_summary$SDRain,
                      na.rm = TRUE) * 1.1)) +
  labs(y = "Total monthly rainfall (mm)", x = NULL, fill = "Population") +
  theme_fig1 +
  theme(legend.position = "none")
```

## Panel E — Activity timeline

```{r fig1-panelE}
# Convert day-of-year to decimal month so x-axis matches panels C & D (scale 1–12)
d2m <- function(day) day * 12 / 365

create_daily_activity <- function(entry_day, emergence_day, site_name) {
  data.frame(
    Site   = site_name,
    Month  = d2m(1:365),
    Status = ifelse(1:365 >= entry_day & 1:365 <= emergence_day,
                    "Hibernation", "Active"))
}

daily_activity <- rbind(
  create_daily_activity(152, 247, "OR"),   # Jun 1 – Sep 4
  create_daily_activity(134, 276, "CP"))   # May 14 – Oct 3

hiber_timing <- data.frame(
  Site     = c("OR", "OR", "CP", "CP"),
  Event    = c("Entry", "Emergence", "Entry", "Emergence"),
  Month    = d2m(c(152, 247, 134, 276)),
  MinMonth = d2m(c(129, 215, 105, 250)),
  MaxMonth = d2m(c(173, 267, 154, 295)),
  y        = c(2, 2, 1, 1))

activity_plot <- ggplot(daily_activity,
                        aes(x = Month, y = Site,
                            fill = interaction(Site, Status))) +
  geom_tile(color = NA, height = 0.8, alpha = 0.4) +
  geom_errorbar(data = hiber_timing,
                aes(xmin = MinMonth, xmax = MaxMonth, y = y, color = Site),
                width = 0.4, linewidth = 1.2,
                orientation = "y", inherit.aes = FALSE) +
  geom_point(data = hiber_timing,
             aes(x = Month, y = y, color = Site),
             size = 3, inherit.aes = FALSE) +
  scale_fill_manual(
    values = c("OR.Active"          = "white",  "OR.Hibernation"  = "coral1",
               "CP.Active"          = "white",  "CP.Hibernation"  = "cornflowerblue"),
    guide  = "none") +
  scale_color_manual(values = site_colors, guide = "none") +
  scale_x_continuous(breaks = 1:12, labels = month_letters,
                   limits = c(0.5, 12.5), expand = expansion(add = 0)) +
  scale_y_discrete(labels = function(x) "0000") +
  labs(x = NULL, y = "Dormant season") +
  theme_fig1 +
  theme(panel.grid.major.x = element_line(color = "gray90", linetype = "dotted"),
        axis.text.y = element_text(color = "white"))
```

## Assemble Figure 1

```{r fig1-assemble, fig.width=7, fig.height=6.5}

# Shared legend from temp plot
legend_fig1 <- get_legend(
  temp_plot +
    theme(legend.position      = "bottom",
          legend.justification = "center",
          legend.direction     = "horizontal"))

# Left column: species photo + map
# Extra padding on each plot so labels sit above, not on top of, the content
context_col <- plot_grid(
  p_species + theme(plot.margin = margin(16, 8, 5, 8)),
  map1,
  ncol        = 1,
  labels      = c("A", "B"),
  label_size  = 14,
  label_x     = 0,
  label_y     = 1,
  hjust       = -0.2,
  vjust       = 1.2,
  rel_heights = c(0.85, 1.15))

# Right column: climate + activity, all with matching top margin
climate_and_activity <- plot_grid(
  temp_plot     + theme(plot.margin = margin(24, 8, 5, 14)),
  rain_plot     + theme(plot.margin = margin(24, 8, 5, 14)),
  activity_plot + theme(plot.margin = margin(24, 8, 5, 14)),
  ncol        = 1,
  labels      = c("C", "D", "E"),
  label_size  = 14,
  label_x     = 0,
  label_y     = 1,
  hjust       = -0.2,
  vjust       = 1.2,
  align       = "hv",
  axis        = "tblr",   
  rel_heights = c(1, 1, 0.7))

# Add shared legend below right column
data_col <- plot_grid(
  climate_and_activity,
  legend_fig1,
  ncol        = 1,
  rel_heights = c(1, 0.07))

# Final figure
figure1 <- plot_grid(context_col, data_col,
                     ncol       = 2,
                     rel_widths = c(1, 1.2))

figure1

ggsave(here("figures", "Figure1_StudySystem.jpg"),
       figure1, width = 7, height = 6.5, units = "in", dpi = 300)

ggsave(here("figures", "Figure1_StudySystem.pdf"),
       figure1, width = 7, height = 6.5, units = "in")
```

---

# Figure 2 — Senescence Trajectories (4-panel)

**Required files:**

- `bafta_outputs/out_CM_CP.rds`
- `bafta_outputs/out_CM_OR.rds`
- `basta_outputs/out_CP_GM.rds`
- `basta_outputs/out_OR_WE.rds`
- `data/growth_data.csv`
- `data/reproductive_data5.csv`

```{r figure2-data}

# ── Shared aesthetics ─────────────────────────────────────────────────────────
site_colors2 <- c("Lowland" = "coral1", "Highland" = "cornflowerblue")

theme_publication <- theme_classic(base_size = 11) +
  theme(axis.text   = element_text(color = "black"),
        axis.line   = element_line(color = "black"),
        plot.margin = margin(12, 8, 5, 8),
        legend.text  = element_text(size = 9),
        legend.title = element_blank())

# ── Load BaFTA outputs ────────────────────────────────────────────────────────
out_CM_CP <- readRDS(here("bafta_outputs", "out_CM_CP.rds"))
out_CM_OR <- readRDS(here("bafta_outputs", "out_CM_OR.rds"))

# ── Load BaSTA outputs ────────────────────────────────────────────────────────
out_CP_GM    <- readRDS(here("basta_outputs", "out_CP.rds"))      # Highland
out_OR_WE <- readRDS(here("basta_outputs", "out_OR_WE.rds"))   # Lowland

# ── Load main dataset ─────────────────────────────────────────────────────────
data <- read.csv(here("data", "processed", "reproductive_data.csv"),
                 na.strings = c("", "NA", ".", "?")) %>%
  dplyr::select(Year, FemID, Site, Age, Clutch) %>%
  dplyr::rename(year = Year, indID = FemID, site = Site, nOffspring = Clutch) %>%
  dplyr::mutate(indID      = as.factor(indID),
                site       = as.factor(site),
                Age        = as.integer(Age),
                nOffspring = as.integer(nOffspring)) %>%
  dplyr::filter(!is.na(indID), indID != 0,
                !is.na(Age),   Age  != 0,
                !is.na(nOffspring), nOffspring != 0)
```

## Panel A — Growth curves

```{r fig2-panelA, message=FALSE}
growth_data_all <- read.csv(here("data", "processed", "growth_data.csv"),
                            na.strings = c("", ",", "NA", "."),
                            header = TRUE)

recode_site <- c(OR = "Lowland", CP = "Highland")

# ── Adults (age ≥ 2): von Bertalanffy via nlme ───────────────────────────────
growth_adults <- growth_data_all %>%
  filter(!is.na(FSVLAFTER), !is.na(SITE.), !is.na(Age),
         !is.na(FEMID.), Age >= 2)

fit_vb_OR <- nlme(
  FSVLAFTER ~ Linf * (1 - exp(-K * (Age - t0))),
  data    = growth_adults %>% filter(SITE. == "OR"),
  fixed   = Linf + K + t0 ~ 1,
  random  = Linf ~ 1 | FEMID.,
  start   = c(Linf = 74, K = 0.044, t0 = 0),
  control = nlmeControl(maxIter = 200, msMaxIter = 200))

fit_vb_CP <- nlme(
  FSVLAFTER ~ Linf * (1 - exp(-K * (Age - t0))),
  data    = growth_adults %>% filter(SITE. == "CP"),
  fixed   = Linf + K + t0 ~ 1,
  random  = Linf ~ 1 | FEMID.,
  start   = c(Linf = 89, K = 0.044, t0 = 0),
  control = nlmeControl(maxIter = 200, msMaxIter = 200))

# Population-level predictions (level = 0)
newdata_ages <- data.frame(Age = seq(2, 14, length.out = 100))

pred_OR <- newdata_ages %>%
  mutate(FSVLAFTER = predict(fit_vb_OR, newdata = newdata_ages, level = 0),
         Site = "OR")

pred_CP <- newdata_ages %>%
  mutate(FSVLAFTER = predict(fit_vb_CP, newdata = newdata_ages, level = 0),
         Site = "CP")

# Bootstrap 95% CIs (parametric, from fixed-effect vcov)
set.seed(123)

boot_vb <- function(fit, newages, n = 1000) {
  beta  <- fixef(fit)
  sims  <- MASS::mvrnorm(n, beta, vcov(fit))
  preds <- apply(sims, 1, function(b) {
    b["Linf"] * (1 - exp(-b["K"] * (newages - b["t0"])))
  })
  list(lower = apply(preds, 1, quantile, 0.025),
       upper = apply(preds, 1, quantile, 0.975))
}

ci_OR <- boot_vb(fit_vb_OR, newdata_ages$Age)
ci_CP <- boot_vb(fit_vb_CP, newdata_ages$Age)

pred_OR <- pred_OR %>% mutate(lower = ci_OR$lower, upper = ci_OR$upper)
pred_CP <- pred_CP %>% mutate(lower = ci_CP$lower, upper = ci_CP$upper)

pred_all <- bind_rows(pred_OR, pred_CP) %>%
  mutate(site = recode(Site, !!!recode_site))

# ── Early growth (age 0–2): lm ─────────────────────────────────────

# Juvenile means ± SE (ages 0, 0.75) 

juvenile_means <- growth_data_all %>%
  filter(!is.na(FSVLAFTER), !is.na(SITE.), !is.na(Age), Age <= 2, Age != 2) %>%
  mutate(site = recode(SITE., !!!recode_site)) %>%
  group_by(Age, site) %>%
  summarise(
    mean_SVL = mean(FSVLAFTER, na.rm = TRUE),
    sd_SVL   = sd(FSVLAFTER,  na.rm = TRUE),
    .groups  = "drop")

growth_plot_data <- growth_data_all %>%
  mutate(site = recode(SITE., !!!recode_site))

Linf_OR <- fixef(fit_vb_OR)["Linf"]
Linf_CP <- fixef(fit_vb_CP)["Linf"]

# ── Plot ──────────────────────────────────────────────────────────────────────
p_growth <- ggplot() +
  geom_jitter(data = growth_plot_data,
              aes(x = Age, y = FSVLAFTER, colour = site),
              width = 0.15, height = 0, alpha = 0.3, size = 0.5) +
  geom_ribbon(data = pred_all,
              aes(x = Age, ymin = lower, ymax = upper, fill = site),
              alpha = 0.2) +
  geom_line(data = pred_all,
            aes(x = Age, y = FSVLAFTER, colour = site), linewidth = 1.2) +
  geom_hline(yintercept = Linf_OR,
             colour = "coral1",         linetype = "dashed", linewidth = 0.6) +
  geom_hline(yintercept = Linf_CP,
             colour = "cornflowerblue", linetype = "dashed", linewidth = 0.6) +
  geom_pointrange(data = juvenile_means,
                  aes(x = Age, y = mean_SVL,
                      ymin = mean_SVL - sd_SVL,
                      ymax = mean_SVL + sd_SVL,
                      fill = site,
                      group = site),
                  colour = "black",
                  shape = 21,
                  size = 1.2,
                  linewidth = 0.8,
                  position = position_dodge(width = 0.3))+
  scale_x_continuous(breaks = seq(0, 12, by = 2), limits = c(-0.3, 12))+
  scale_colour_manual(values = site_colors2) +
  scale_fill_manual(values   = site_colors2) +
  labs(x = "Age (years)", y = "Snout-vent length (mm)") +
  theme_publication +
  theme(legend.position = "none")

p_growth


```

## Panel B — Cumulative offspring

```{r fig2-panelB}
cumulative_data <- data %>%
  arrange(indID, Age) %>%
  group_by(indID, site) %>%
  mutate(cumulative_offspring = cumsum(nOffspring),
         reproductive_event  = row_number()) %>%
  ungroup() %>%
  filter(Age >= 2, Age <= 12) %>%
  mutate(site = recode(as.character(site), !!!recode_site))

cumulative_summary <- cumulative_data %>%
  group_by(site, Age) %>%
  summarise(
    mean_cumulative = mean(cumulative_offspring),
    se        = sd(cumulative_offspring) / sqrt(n()),
    lower_ci  = mean_cumulative - 1.96 * se,
    upper_ci  = mean_cumulative + 1.96 * se,
    .groups   = "drop")

p_cumulative <- ggplot() +
  geom_line(data = cumulative_data,
            aes(x = Age, y = cumulative_offspring,
                group = indID, color = site),
            alpha = 0.2, linewidth = 0.3) +
  geom_ribbon(data = cumulative_summary,
              aes(x = Age, ymin = lower_ci, ymax = upper_ci, fill = site),
              alpha = 0.3) +
  geom_line(data = cumulative_summary,
            aes(x = Age, y = mean_cumulative, color = site),
            linewidth = 1.5) +
  scale_color_manual(values = site_colors2) +
  scale_fill_manual(values  = site_colors2) +
  scale_x_continuous(breaks = seq(2, 12, 2)) +
  scale_y_continuous(breaks = seq(0, 55, 10)) +
  labs(x = "Age (years)", y = "Cumulative offspring") +
  theme_publication +
  theme(legend.position = "none")
```

## Panel C — Fertility trajectories (BaFTA)

```{r fig2-panelC}
fert_data <- bind_rows(
  data.frame(age    = out_CM_OR$x,
             median = out_CM_OR$fert[, "Mean"],
             lower  = out_CM_OR$fert[, "Lower"],
             upper  = out_CM_OR$fert[, "Upper"],
             site   = "Lowland"),
  data.frame(age    = out_CM_CP$x,
             median = out_CM_CP$fert[, "Mean"],
             lower  = out_CM_CP$fert[, "Lower"],
             upper  = out_CM_CP$fert[, "Upper"],
             site   = "Highland"))

p_fertility <- ggplot(fert_data,
                      aes(x = age, y = median, color = site, fill = site)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, color = NA) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = site_colors2) +
  scale_fill_manual(values  = site_colors2) +
  scale_x_continuous(breaks = seq(2, 12, 2), limits = c(2, 12)) +
  labs(y = "Expected litter size", x = "Age (years)") +
  theme_publication +
  theme(legend.position = "none")
```

## Panel D — Mortality hazard (BaSTA)

```{r fig2-panelD}
# Extract Gompertz-Makeham trajectories from MCMC posteriors (CP highland)
calculate_hazard_GM <- function(basta_out, site_name, ages = 2:12) {
  params <- as.data.frame(basta_out$params)
  lapply(seq_len(nrow(params)), function(i) {
    data.frame(
      age    = ages,
      hazard = exp(params$b0[i] + params$b1[i] * ages) + params$c[i],
      site   = site_name)
  }) %>%
    bind_rows() %>%
    group_by(age, site) %>%
    summarise(
      median = median(hazard),
      lower  = quantile(hazard, 0.025),
      upper  = quantile(hazard, 0.975),
      .groups = "drop")
}

# Extract Weibull-simple trajectories from MCMC posteriors (OR lowland)
# Weibull hazard: mu(x) = b0 * b1 * x^(b1 - 1)
calculate_hazard_WE <- function(basta_out, site_name, ages = 2:12) {
  params <- as.data.frame(basta_out$params)
  lapply(seq_len(nrow(params)), function(i) {
    data.frame(
      age    = ages,
      hazard = params$b0[i] * params$b1[i] * ages^(params$b1[i] - 1),
      site   = site_name)
  }) %>%
    bind_rows() %>%
    group_by(age, site) %>%
    summarise(
      median = median(hazard),
      lower  = quantile(hazard, 0.025),
      upper  = quantile(hazard, 0.975),
      .groups = "drop")
}

haz_OR <- calculate_hazard_WE(out_OR_WE, "Lowland")
haz_CP <- calculate_hazard_GM(out_CP_GM, "Highland")
haz_data <- bind_rows(haz_OR, haz_CP)

p_mortality <- ggplot(haz_data,
                      aes(x = age, y = median, color = site, fill = site)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, color = NA) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = site_colors2) +
  scale_fill_manual(values  = site_colors2) +
  scale_x_continuous(breaks = seq(2, 12, 2), limits = c(2, 12)) +
  labs(x = "Age (years)", y = "Mortality hazard μ(x)") +
  theme_publication +
  theme(legend.position = "none")

```

## Assemble Figure 2

```{r fig2-assemble, fig.width=10, fig.height=8}

# Shared legend from fertility panel
legend_fig2 <- get_legend(
  p_fertility +
    theme(legend.position      = "bottom",
          legend.justification = "center",
          legend.direction     = "horizontal"))

# 2 × 2 grid
panels_2x2 <- plot_grid(
  p_growth, p_cumulative,
  p_fertility, p_mortality,
  ncol       = 2,
  nrow       = 2,
  labels     = c("A", "B", "C", "D"),
  label_size = 14,
  label_x    = 0,
  label_y    = 1,
  hjust      = -0.2,
  vjust      = 1.2,
  align      = "hv",
  axis       = "tblr")

# Attach shared legend
figure2 <- plot_grid(panels_2x2, legend_fig2,
                     ncol        = 1,
                     rel_heights = c(1, 0.06))

figure2

ggsave(here("figures", "Figure2_Senescence.png"),
       figure2, width = 10, height = 8, units = "in", dpi = 600)

ggsave(here("figures", "Figure2_Senescence.pdf"),
       figure2, width = 10, height = 8, units = "in")
```

---

# Session Info

```{r session-info}
sessionInfo()
```
